---
title: "Embedding codeagent as a backend (Contract v1)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Embedding codeagent as a backend (Contract v1)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

**Language:** English | [简体中文](backend-integration-cn.html)

This vignette describes the **stable public interface** for embedding codeagent
as the engine behind a host application (a Shiny app, an API service, another
package). The host keeps full ownership of its **UI, tools, skill content, data,
and provider credentials**; codeagent provides the **agent loop, streaming,
context compaction, permission gate, and skill loading**.

Everything documented here is **Backend Contract v1**. See
[Versioning](#versioning) for the stability promise. A runnable reference lives
at `system.file("examples/backend_integration_demo.R", package = "codeagent")`.

## The boundary

| codeagent provides | The host owns |
|---|---|
| agent loop, multi-turn tool-calling | its UI |
| provider abstraction (any ellmer `Chat`) | its `Chat` (provider / model / key) |
| streaming + typed callbacks | rendering |
| context compaction | domain tools |
| central permission gate | skill content + data |
| skill loading | session storage (optional) |

## 1. Entry: a harness-only client

Pass your own `ellmer::Chat` and set `register_tools = FALSE` so **none** of
codeagent's coding tools (Bash / Write / Edit / Glob / Grep / git / web) are
attached. You get the client state and harness pipeline without registered
tools or an installed permission gate.

```{r}
chat <- ellmer::chat_openai_compatible(
  base_url    = Sys.getenv("MY_BASE_URL"),
  model       = Sys.getenv("MY_MODEL"),
  credentials = function() Sys.getenv("MY_API_KEY")
)

client <- codeagent::codeagent_client(
  chat            = chat,
  register_tools  = FALSE,
  permission_mode = "default",   # default | plan | accept_edits | bypass | ...
  cwd             = getwd()
)
```

`codeagent_client()` returns a `CodeagentClient` with `$chat`, `$settings`, and
`$data_shield` (`NULL` unless enabled). If a Data Shield is enabled on a
harness-only client, attach the host tools first and then install the returned
shield on the chat. For multi-user Shiny apps, create the client inside the
server session (for example via `codeagent_app(client_factory = )`); never share
one mutable client across browser sessions.

## 2. Driving a turn + the callback contract

Use `codeagent_stream()` (blocking; pumps its own event loop) or
`codeagent_stream_async()` (returns a promise). Rendering happens entirely
through **typed callbacks** -- codeagent does not touch your UI.

```{r}
result <- codeagent::codeagent_stream(
  client, user_input,
  on_delta        = function(text_chunk)       { ... }, # assistant text
  on_thinking     = function(chunk)            { ... }, # thinking content
  on_tool_request = function(x)                { ... }, # id, name, arguments, intent
  on_tool_result  = function(x)                { ... }, # six fields; see below
  on_error        = function(message, recovered) { ... },
  on_usage        = function(usage)            { ... }, # end-of-turn usage
  on_tick         = function()                 { ... }  # sync only; ~100 ms heartbeat
)
# returns invisibly: list(text, usage, stop_reason, finish_reason)
```

Callback payloads:

| Callback | Argument |
|---|---|
| `on_delta` | `text_chunk` (character) |
| `on_thinking` | thinking chunk |
| `on_tool_request` | `list(id, name, arguments, intent)` -- fires **before** the gate |
| `on_tool_result` | `list(id, name, display, value, is_error, artifact)` in that exact order; `artifact` is appended after the five original fields |
| `on_error` | `message, recovered` |
| `on_usage` | usage object, including `n_tokens`, `model_limit`, `warning_state`, and `cost_last` |
| `on_tick` | no arguments; available on synchronous `codeagent_stream()` only |

The asynchronous function resolves to the same four-field result. Provider
finish reasons are normalized into `stop_reason`; the original mapped provider
value is retained as `finish_reason`.

## 3. Rich tool results (text / table / image / code / diff / error)

A tool's `value` is the portable text the model sees and the final fallback for
all UIs. To also expose a **rich, UI-neutral artifact**, return `tool_result()`.
The result carries three deliberately separate channels:

- `artifact`: the primary cross-UI protocol (`schema`, `version`, `kind`,
  `status`, `icon`, `title`, `payload`);
- `display`: an optional official shinychat adapter; non-shinychat hosts should
  not parse or depend on it;
- `value`: portable text for the model and unsupported/malformed artifacts.

```{r}
my_tool <- ellmer::tool(
  function(name) {
    df <- summarise_something(name)
    codeagent::tool_result(
      sprintf("%d x %d summary", nrow(df), ncol(df)),
      kind    = "table",
      payload = list(df = df),
      title   = "Summary"
    )
  },
  name = "Summarise",
  description = "Summarise a named dataset.",
  arguments = list(
    name = ellmer::type_string("Dataset name.")
  )
)
```

The version-1 artifact is available as `extra$codeagent$artifact` on a result and
as `artifact` on the stream event. Hosts should use
`tool_result_artifact(event_or_result)` rather than reaching into either
structure directly. It validates the schema and outer shape and, by default,
accepts version 1 only. It returns `NULL` for unknown or malformed artifacts,
allowing a safe fallback through `tool_result_value(event_or_result)`.

| `kind` | Typical `payload` |
|---|---|
| `text` | `list(text = )` |
| `table` | `list(df = <data.frame>)` |
| `image` | `list(images = list(list(mime = , b64 = )), output = )` |
| `code` | `list(text = , lang = , filename = , output = )` |
| `diff` | `list(old = , new = , path = )` |
| `error` | `list(message = , detail = )` |

A non-Shiny host renders the artifact itself, for example:

```{r}
on_tool_result <- function(event) {
  artifact <- codeagent::tool_result_artifact(event)
  if (!is.null(artifact) && identical(artifact$kind, "table")) {
    my_render_table(artifact$payload$df)
  } else {
    render_plain_text(codeagent::tool_result_value(event))
  }
}
```

See `vignette("tool-artifacts")` for version negotiation, failure behavior, and
the browser trust boundary.

## 4. Host tools + the permission gate

Register your tools the standard ellmer way, then **declare each tool's
capability** so the central gate governs it like a native tool.

```{r}
chat$register_tool(my_tool)
codeagent::register_tool_meta("Summarise", capability = "read") # read|write|exec|net
```

On a harness-only client (`register_tools = FALSE`) the gate is **not** installed
automatically. Install it after attaching your tools so approvals route to your
`ask_fn`. The public `tools` argument is the standalone equivalent of
`settings$tools`:

```{r}
tool_policy <- list(
  overrides    = list(Summarise = "allow"),       # allow | ask | deny
  capabilities = list(exec = "ask", net = "deny")
)

codeagent::install_permission_gate(
  chat,
  permission_mode = "default",
  tools = tool_policy,
  # Alternatively classify tools here instead of register_tool_meta():
  tool_meta = list(Summarise = "read"),
  ask_fn = function(name, input, id = NULL) {
    host_request_approval(id, name, input) # logical or promise<logical>
  }
)
```

The optional `id` is the tool-call ID and matches `on_tool_request$id`.
`install_permission_gate()` is idempotent per chat: another call refreshes its
live mode, approval callback, and policy rather than stacking another gate.

> **Important:** an *undeclared* tool defaults to capability `"read"` and is
> allowed without sensitive-operation gating. If your tool executes code,
> writes files, or accesses the network, declare it as `"exec"`, `"write"`, or
> `"net"` so the gate can ask or deny it. Built-in metadata remains
authoritative and cannot be downgraded by host registration.

Policy precedence is per-tool `overrides`, then per-capability policy, then the
active permission mode and rules. The complete policy shape is
`sets` / `capabilities` / `overrides`; pass it as `tools =` to the standalone
installer or configure it as `settings$tools` on a full codeagent client.

## 5. Skills

Place host skills in a supported project-local skill directory using the
`<name>/SKILL.md` format (for example,
`.btw/skills/my_skill/SKILL.md`). Skill *content* is yours; codeagent only
discovers, loads, and injects it.

```{r}
skills <- codeagent::list_skills_meta(cwd = getwd())
prompt <- codeagent::load_skill_prompt(
  "my_skill", args = "optional arguments", cwd = getwd()
)
hint <- codeagent::build_skill_hint(cwd = getwd(), max_tokens = 1000L)
```

## 6. Provider

`codeagent_client(chat = )` accepts any `ellmer::Chat` (OpenAI-compatible,
Databricks, Anthropic, Gemini, Bedrock, Azure, and others). The lower-level
stream functions also accept a bare `ellmer::Chat`; list-based `$stream_async`
duck typing exists for tests, not as the client-construction contract. The host
owns and supplies credentials; codeagent uses the supplied Chat rather than
owning those credentials.

## 7. Versioning {#versioning}

Backend Contract **v1** is the surface below, with the signatures and behavior
documented above. Changes follow semantic versioning; breaking changes bump the
major and are announced in `NEWS.md`. The guard test
`test-backend-contract.R` fails if the exported surface drifts.

- `codeagent_client(register_tools = FALSE)` -> `{chat, settings, data_shield}`
- `codeagent_app(client_factory = )` per-session client contract
- `codeagent_stream()` / `codeagent_stream_async()` + callbacks +
  `list(text, usage, stop_reason, finish_reason)`
- `agent_loop()`
- `tool_result()`
- `tool_result_artifact()` / `tool_result_value()`
- `register_tool_meta()`
- `install_permission_gate()` (+ `ask_fn(name, input, id = NULL)` contract)
- `DataShield` / `shield_describe()` / `shield_egress()` / `shield_regex()` /
  `shield_ingress()` / `shield_tool_policy()` / `shield_sandbox()` /
  `shield_reviewer()`
- `list_skills_meta()` / `load_skill_prompt()` / `build_skill_hint()`
- `CompactionController` (compaction runs automatically; this is the injectable controller)
- `switch_model()`
- `settings$tools` policy: `sets` / `capabilities` / `overrides`
