---
title: "Embedding codeagent as a backend (Contract v1)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Embedding codeagent as a backend (Contract v1)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette describes the **stable public interface** for embedding codeagent
as the engine behind a host application (a Shiny app, an API service, another
package). The host keeps full ownership of its **UI, tools, skill content, data,
and provider credentials**; codeagent provides the **agent loop, streaming,
context compaction, permission gate, and skill loading**.

Everything documented here is **Backend Contract v1**. See
[Versioning](#versioning) for the stability promise. A runnable reference lives
at `system.file("examples/backend_integration_demo.R", package = "codeagent")`.

## The boundary

| codeagent provides | The host owns |
|---|---|
| agent loop, multi-turn tool-calling | its UI |
| provider abstraction (any ellmer `Chat`) | its `Chat` (provider / model / key) |
| streaming + typed callbacks | rendering |
| context compaction | domain tools |
| central permission gate | skill content + data |
| skill loading | session storage (optional) |

## 1. Entry: a harness-only client

Pass your own `ellmer::Chat` and set `register_tools = FALSE` so **none** of
codeagent's coding tools (Bash / Write / Edit / Glob / Grep / git / web) are
attached — you get only the harness.

```{r}
chat <- ellmer::chat_openai_compatible(
  base_url    = Sys.getenv("MY_BASE_URL"),
  model       = Sys.getenv("MY_MODEL"),
  credentials = function() Sys.getenv("MY_API_KEY")
)

client <- codeagent::codeagent_client(
  chat            = chat,
  register_tools  = FALSE,
  permission_mode = "default",   # default | plan | accept_edits | bypass | ...
  cwd             = getwd()
)
```

`codeagent_client()` returns a `CodeagentClient` with `$chat`, `$settings`, and
`$data_shield` (`NULL` unless enabled). For multi-user Shiny apps, create the
client inside the server session (for example via `codeagent_app(client_factory=)`),
never share one mutable client across browser sessions.

## 2. Driving a turn + the callback contract

Use `codeagent_stream()` (blocking; pumps its own event loop) or
`codeagent_stream_async()` (returns a promise). Rendering happens entirely
through **typed callbacks** — codeagent does not touch your UI.

```{r}
codeagent::codeagent_stream(
  client, user_input,
  on_delta        = function(text_chunk) { ... },  # incremental assistant text
  on_thinking     = function(chunk)      { ... },  # reasoning / thinking content
  on_tool_request = function(x)          { ... },  # list(id, name, arguments, intent)
  on_tool_result  = function(x)          { ... },  # list(id, name, display, value, is_error)
  on_usage        = function(usage)      { ... },  # token usage
  on_tick         = function()           { ... }   # ~100 ms heartbeat (spinners)
)
# returns invisibly: list(text, usage, stop_reason)
```

Callback payloads:

| Callback | Argument |
|---|---|
| `on_delta` | `text_chunk` (character) |
| `on_thinking` | thinking chunk |
| `on_tool_request` | `list(id, name, arguments, intent)` — fires **before** the gate |
| `on_tool_result` | `list(id, name, display, value, is_error)` |
| `on_usage` | usage object |
| `on_tick` | none |

## 3. Rich tool results (text / table / image / error)

A tool's `value` is the text the model sees. To also render a **rich artifact**
in your UI, return `tool_result()`, which attaches a typed display card
delivered as `on_tool_result$display`.

```{r}
my_tool <- ellmer::tool(
  function(name) {
    df <- summarise_something(name)
    codeagent::tool_result(
      sprintf("%d x %d summary", nrow(df), ncol(df)),
      kind    = "table",
      payload = list(df = df),
      title   = "Summary"
    )
  },
  name = "Summarise", description = "...", arguments = list(...)
)
```

`display$toolcard$kind` + `display$toolcard$payload` carry the structured
artifact for any host to render; codeagent's own Shiny app additionally receives
a pre-rendered `display$html` / `display$right_output`.

| `kind` | `payload` |
|---|---|
| `text` | `list(text=)` |
| `table` | `list(df = <data.frame>)` |
| `image` | `list(images = list(list(mime=, b64=)), output=)` |
| `code` | `list(text=, lang=, filename=, output=)` |
| `error` | `list(message=, detail=)` |

A non-Shiny host renders the artifact itself, e.g.:

```{r}
on_tool_result <- function(x) {
  if (identical(x$display$toolcard$kind, "table"))
    my_render_table(x$display$toolcard$payload$df)
}
```

## 4. Host tools + the permission gate

Register your tools the standard ellmer way, then **declare each tool's
capability** so the central gate governs it like a native tool.

```{r}
chat$register_tool(my_tool)
codeagent::register_tool_meta("RunAnalysis", capability = "exec")  # read|write|exec|net
```

On a harness-only client (`register_tools = FALSE`) the gate is **not** installed
automatically — install it once, after attaching your tools, so they are
governed and approvals route to your `ask_fn`:

```{r}
codeagent::install_permission_gate(
  chat,
  permission_mode = "default",
  tool_meta = list(RunAnalysis = "exec"),      # optional: declare capabilities here
  ask_fn = function(name, input, id = NULL) {   # `id` = tool-call id, matches on_tool_request
    host_request_approval(id, name, input)       # return a logical or a promise<logical>
  }
)
```

> **Important:** an *undeclared* tool defaults to capability `"read"` and is
> allowed **without gating**. If your tool executes code, writes files, or hits
> the network, declare it (`"exec"`/`"write"`/`"net"`) so the gate can `ask` or
> `deny` it. Built-in tool metadata stays authoritative.

Fine-grained control is available through `settings$tools`:

```{r}
settings$tools <- list(
  overrides    = list(RunAnalysis = "ask"),      # per-tool: allow | ask | deny
  capabilities = list(exec = "ask", net = "deny")# per-capability policy
)
```

## 5. Skills

Point codeagent at your own `<name>/SKILL.md` directories (scanned under `cwd`).
Skill *content* is yours; codeagent only loads and injects it.

```{r}
codeagent::list_skills_meta(cwd = getwd())
codeagent::load_skill_prompt("my_skill", cwd = getwd())
codeagent::build_skill_hint(...)
```

## 6. Provider

`chat` accepts any `ellmer::Chat` (OpenAI-compatible, Databricks, Anthropic,
Gemini, Bedrock, Azure, …) — or any object exposing `$stream_async()`. The host
owns and supplies credentials; codeagent never reads them.

## 7. Versioning {#versioning}

Backend Contract **v1** = the symbols below, with the signatures documented
above. Changes follow semantic versioning; breaking changes bump the major and
are announced in `NEWS.md`. The guard test `test-backend-contract.R` fails if
this surface drifts.

- `codeagent_client(register_tools = FALSE)` → `{chat, settings, data_shield}`
- `codeagent_app(client_factory = )` per-session client contract
- `codeagent_stream()` / `codeagent_stream_async()` + callbacks + `list(text, usage, stop_reason)`
- `agent_loop()`
- `tool_result()`
- `register_tool_meta()`
- `install_permission_gate()` (+ `ask_fn(name, input, id = NULL)` contract)
- `DataShield` / `shield_describe()` / `shield_egress()` / `shield_regex()` /
  `shield_ingress()`
- `list_skills_meta()` / `load_skill_prompt()` / `build_skill_hint()`
- `CompactionController` (compaction runs automatically; this is the injectable controller)
- `switch_model()`
- `settings$tools` policy: `sets` / `capabilities` / `overrides`
