---
title: "Context management and compaction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Context management and compaction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, purl=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

**Language:** English | [简体中文](compaction-cn.html)

codeagent keeps long conversations within the model's context window with
turn-boundary compaction, default-on mid-loop tool-result snipping, and reactive
prompt-too-long recovery. None of these accounting paths performs an implicit
remote token-count request.

## Dynamic context window

The raw context window is resolved in this order:

1. A valid positive `CODEAGENT_MAX_CONTEXT_TOKENS` value.
2. A `[1m]` suffix in the model name, which selects 1,000,000 tokens.
3. A best-effort provider-reported capability when present, otherwise a built-in
   longest-substring table for Claude, GPT, Gemini, DeepSeek, Llama, Qwen, and
   Mistral models. The resulting capability is accepted only when it is at least
   100,000 tokens; otherwise resolution uses the default.
4. The 200,000-token default.

`CODEAGENT_MODEL_LIMIT`, when set, directly replaces `settings$model_limit`
after settings are loaded. This is the value passed to turn-boundary and the
usual mid-loop trigger calculation.

For context-left status and the dynamic auto threshold, the effective window is
`raw window - output reserve`; the reserve is model-specific and capped at
20,000 tokens. `CODEAGENT_AUTO_COMPACT_WINDOW` can cap the raw window used by
that effective-window calculation. The dynamic threshold then subtracts a
further 13,000-token buffer.

```{r, purl=FALSE}
# Turn-boundary compaction uses settings$model_limit - 33,000 directly.
# With the default 200,000-token model_limit, it triggers at about 167,000.
```

## Token accounting

`token_count_with_estimation(chat, allow_network = FALSE)` first uses the last
recorded API usage, including cached input when reported. If no positive usage
is available, it estimates from local conversation text at roughly 3.5
characters per token. Only an explicit call with `allow_network = TRUE` may use
`chat$token_count(include = "complete")`.

## Five mechanisms and the automatic summary flow

The implementation has five mechanisms:

1. **L1 `snip_old_tools()`** replaces eligible old tool results with a small
   placeholder, without an LLM call.
2. **L2 `session_memory_compact()`** incrementally summarizes early turns.
3. **L3 `full_compact()`** creates one structured nine-section summary and
   replaces history with that summary, retaining the latest plain user turn
   when safe.
4. **L4 `ptl_fallback()`** drops oldest turns after a 413/prompt-too-long error.
5. **L5 `context_collapse()`** is a utility that truncates large tool-result
   values in place; it is not part of the normal automatic summary chain.

`CompactionController$compact_now()` runs L1, then tries L2. It runs L3 only
when L2 could not run, such as when there are too few suitable turns. Three
consecutive failed compactions open the circuit breaker; a successful compact
resets its failure count.

## Exact timing and defaults

```
token count = last recorded API usage (including cached input)
              else local char/3.5 estimate

TURN BOUNDARY -- .turn_setup(), before each chat$chat()
  CompactionController$maybe_compact(
    model_limit = settings$model_limit, compact_model = resolved fast model)

  enabled AND failures < 3 AND tokens >= model_limit - 33,000 ?
    yes -> compact_now()
             L1 snip_old_tools()
             L2 session_memory_compact()
                or, only if L2 did not run, L3 full_compact()

BETWEEN TOOL ROUNDS -- ellmer on_tool_result callback
  settings$midloop_compact = TRUE (default)
  AND tokens >= midloop trigger ?

    default path:
      snip_old_tools(keep_recent_turns = 10,
                     target_tokens = dynamic auto threshold / 2)
      # budget-aware micro-snip; no LLM call

    settings$midloop_full_compact = TRUE (opt-in) path:
      compact_now()
      # blocking L1 -> L2-or-L3 path; used instead of the micro-snip path

REACTIVE -- provider reports 413 / prompt too long
  ptl_fallback(chat, drop_turns = 3L, error_msg = message)
    parsed real limit available -> drop oldest turns until the local estimate
                                   is at most 90% of that limit
    no parsed limit             -> drop the oldest 3 turns
  error recovery then retries the request
```

The default mid-loop trigger is `settings$model_limit - 33,000`. It can be
replaced by positive `settings$midloop_threshold` or
`options(codeagent.midloop_threshold=)`. The micro-snip target and number of
recent turns can similarly be replaced by `settings$midloop_snip_target` /
`options(codeagent.midloop_snip_target=)` and
`settings$midloop_keep_recent` / `options(codeagent.midloop_keep_recent=)`.
Options can also opt in when the corresponding setting is false:
`codeagent.midloop_compact` and `codeagent.midloop_full_compact`.

Mid-loop work currently rides `on_tool_result`, so it runs after a tool result
and before a subsequent model round. It does not run before every provider
request. `on_tool_request` cannot close that timing gap because it fires after
the model request, inside tool invocation; an upstream `on_turn_start` hook
would provide the cleaner timing.

## Context-left indicator

The REPL and Shiny show "N% context left" using
`calculate_token_warning_state()`. With automatic compaction enabled, the
percentage is relative to the dynamic auto threshold, not the full raw window.
Warning, error, compact, and blocking states use separate buffers, and the UI
changes level as those boundaries are crossed.

## Relevant signatures and defaults

```{r, purl=FALSE}
# Exported R6 controller
ctrl <- CompactionController$new()
ctrl$maybe_compact(
  chat,
  model_limit = 200000L,
  compact_model = codeagent:::.HAIKU_MODEL
)
ctrl$compact_now(
  chat,
  compact_model = codeagent:::.HAIKU_MODEL
)
ctrl$handle_ptl_error(chat, error = NULL)
ctrl$reset_failures()
ctrl$failure_count()

# Internal helpers shown here to make their operational defaults explicit
codeagent:::token_count_with_estimation(chat, allow_network = FALSE)
codeagent:::ptl_fallback(chat, drop_turns = 3L, error_msg = NULL)
codeagent:::context_collapse(chat, max_chars = 200L)
```

The actual compaction model is normally selected by the client from
`settings$compact_model`, then `settings$small_fast_model`, and finally the
package fallback; callers do not need to reference the internal fallback
constant.

## Controls

```{r, purl=FALSE}
Sys.setenv(CODEAGENT_DISABLE_COMPACT = "1")     # disable automatic compaction
Sys.setenv(CODEAGENT_MAX_CONTEXT_TOKENS = "500000")  # override raw window
# Manual compaction from the REPL / Shiny, with optional focus instructions:
# /compact
# /compact preserve debugging details and exact file paths
```

Automatic compaction is enabled only when `CODEAGENT_DISABLE_COMPACT` is unset
or empty; any non-empty value disables it. `/compact` is a local slash command,
not a model request, and can carry optional focus instructions.
