---
title: "Context management and compaction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Context management and compaction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, purl=FALSE, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

codeagent keeps long conversations within the model's context window using an
automatic, Claude Code-aligned compaction strategy.

## Dynamic context window

The context window is resolved from the model rather than hard-coded. Resolution
order (highest priority first):

1. `CODEAGENT_MAX_CONTEXT_TOKENS` environment variable.
2. A `[1m]` suffix in the model name (forces a 1,000,000-token window).
3. A built-in capability table (claude / gpt / gemini / deepseek / ...), or a
   provider-reported window when available.
4. A 200,000-token default.

```{r, purl=FALSE, purl=FALSE}
# The effective window reserves output tokens; the auto-compaction threshold
# sits one buffer below that (e.g. ~167K for a 200K Claude model).
```

## When compaction triggers

Compaction runs when the estimated token usage crosses the auto-compaction
threshold (`effective window - output reserve - buffer`). Token usage prefers
the *real* usage from the last API exchange (`Chat$get_tokens()`), falling back
to a character estimate.

## Two-level flow

1. **Snip** old tool results (cheap pre-step).
2. **Session-memory compaction** — an incremental summary of the earliest turns.
3. **Full compaction** — a single 9-section structured summary (verbatim Claude
   Code prompt) replacing the conversation, used when session-memory can't run.

A circuit breaker stops auto-compaction after repeated failures, and a 413 /
prompt-too-long error triggers a reactive fallback that drops the oldest turns
(parsing the real context limit from the error when present).

## Context-left indicator

Both the REPL banner and the Shiny status bar show "N% context left", turning
yellow near the warning line and red near the error/blocking line
(`calculate_token_warning_state()`).

## Controls

```{r, purl=FALSE, purl=FALSE}
Sys.setenv(CODEAGENT_DISABLE_COMPACT = "1")     # disable auto-compaction
Sys.setenv(CODEAGENT_MAX_CONTEXT_TOKENS = "500000")  # override the window
# Manual compaction from the REPL / Shiny:
# /compact
```
