---
title: "Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, purl=FALSE, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Precedence

Settings resolve from lowest to highest priority:

```
package defaults  <  ~/.codeagent/settings.json  <  .codeagent/settings.json  <  environment variables
```

Scaffold a settings file:

```{r, purl=FALSE, purl=FALSE}
use_codeagent_settings(scope = "user")     # ~/.codeagent/settings.json
```

## settings.json

```json
{
  "provider": "openai_compatible",
  "model": "main",
  "env": {
    "CODEAGENT_BASE_URL": "https://YOUR-WORKSPACE.cloud.databricks.net/serving-endpoints",
    "CODEAGENT_MODEL": "your-main-endpoint-name",
    "CODEAGENT_HEAVY_MODEL": "your-heavy-endpoint-name",
    "CODEAGENT_FAST_MODEL": "your-fast-endpoint-name"
  },
  "permissions": { "allow": [], "deny": [], "ask": [], "defaultMode": "default" },
  "sandbox": { "enabled": false, "allow_network": true },
  "effortLevel": "high"
}
```

The `env` block is applied before environment variables are read, so it works
even under `Rscript --vanilla`.

## Environment variables

| Variable | Purpose |
|----------|---------|
| `CODEAGENT_BASE_URL` | OpenAI-compatible endpoint |
| `CODEAGENT_MODEL` | Default model (the `"main"` tier) |
| `CODEAGENT_HEAVY_MODEL` | `"heavy"` tier — most capable model |
| `CODEAGENT_FAST_MODEL` | `"fast"` tier — compaction / classification model |
| `CODEAGENT_API_KEY` | API key (never put this in `settings.json`) |
| `CODEAGENT_MAX_CONTEXT_TOKENS` | Override the context window |
| `CODEAGENT_DISABLE_COMPACT` | Disable auto-compaction |
| `CODEAGENT_PERMISSION_MODE` | Default permission mode |

## API keys

Keep API keys in `~/.Renviron` as `CODEAGENT_API_KEY`, or use the OS keyring
(offered by `use_codeagent_setup()`). Never store keys in `settings.json`.

## Per-project multi-client config

```yaml
# codeagent.md
---
client:
  gpt41:    openai/gpt-4.1
  deepseek: openai/deepseek-r1
btw_groups: [docs, git, pkg]
permission_mode: bypass
---
Follow tidyverse style.
```

```{r, purl=FALSE, purl=FALSE}
client <- codeagent_client_config(alias = "gpt41")
```
