---
title: "Data Shield: strict data-safety mode (design preview)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Shield: strict data-safety mode (design preview)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status — P0/P0.5 available; fuller design in progress.** The **P0
> foundation is wired**: `codeagent_client(data_shield = ...)` and
> `install_data_shield()` wrap tool results through the shape-based egress
> row-cap. P0.5 adds `register_protected_data()` + deterministic `value_match`
> for targeted single-value leaks. Ambient injection stays schema-only. The
> fuller strategy API below (`shield_egress`/`DescribeData`/`reviewer`/`sandbox`)
> is the **target design / roadmap** and not all wired yet. Off by default
> (`data_shield = NULL`).

## Why

When codeagent is embedded as a backend over **sensitive data** (clinical,
financial, PII), the guarantee we want is: the LLM may see **metadata and
descriptive summaries**, but **never raw row-level data**. Data Shield is an
**opt-in, pluggable valve** that enforces this — off by default
(`data_shield = NULL`), composed from independent strategies when on.

## The core: two edges

Strip away the agent machinery and data can reach the LLM through **only two
inbound edges**. Guard both and you have guarded everything:

1. **Prompt / system-prompt content** — including the framework's own
   **ambient-context auto-injection** (codeagent can inject a summary of R
   session objects into the system reminder). This is the part we control and
   must keep **schema-only** (names / types / dimensions), never values.
2. **Tool results** — whatever a tool returns and that is fed back to the model.

Everything else reduces to these two (RAG and errors arrive as one of them;
model-generated content is outbound, not inbound). The guarantee applies
**recursively to sub-agents** — each has its own two edges.

> Blind spot to handle separately: an **image/multimodal** tool result (a
> rendered table/plot of raw rows) bypasses text scanning.

## Design: composable strategies (not a fixed mode)

`data_shield` is `NULL` (off) or a **list of independent strategies** you
compose freely. The execution model borrows llm-guard's scanner pipeline
(ordered, `fail_fast`, each returns `sanitized / valid / score / action`); the
per-strategy `on_fail` action and named registry borrow from guardrails-ai.
All pure R — no Python.

```{r}
# Target API (in progress):
codeagent_client(..., data_shield = list(
  shield_describe(k_anon = 5),                                  # safe metadata only
  shield_egress(detectors = c("value_match", "regex"),          # main boundary
                max_rows = 0, on_fail = "block"),
  shield_ingress(langs = c("R", "python", "bash"), on_fail = "ask"),
  shield_reviewer(model = "<a small, fast model>", scope = "exec"), # small-model audit
  shield_sandbox(no_network = TRUE),                            # OS confinement
  shield_narrow_tools()
))
```

The main boundary is **edge 2 (tool results)**; the rest (sandbox, ingress
blacklist, reviewer, tool narrowing) are **defense-in-depth**, not the boundary.

## P0 — the foundation (available now)

The minimal, deterministic slice that already gives real protection:

```{r}
# Turn it on (register_tools = TRUE path):
client <- codeagent_client(chat, data_shield = list(max_rows = 0))

# Harness-only client: install after attaching your own tools:
client <- codeagent_client(chat, register_tools = FALSE)
chat$register_tool(my_tool)
install_data_shield(chat, max_rows = 0)
```

- **Edge 2 — shape-based egress row-cap.** codeagent does **not** inspect code
  or block `print`. It looks only at the **shape** of a tool's returned text:
  output with a data.frame / tibble print signature or a many-row rectangular
  table is truncated to a shape summary; scalars, messages, **model summaries**,
  plots and errors pass through untouched. Content-agnostic and tunable via
  `max_rows`.
- **Edge 1 — ambient stays schema-only.** codeagent's ambient injection already
  emits only `name [data.frame N x M: col:type, ...]` (no values); Data Shield
  keeps it that way.

### Runtime uploads in Shiny

The dataset does **not** need to be known when the app starts. Register it in
the upload observer immediately after reading it; tools may already be attached
and wrapped, because value matching reads the live index at invocation time.

```{r}
# Per Shiny session:
data_env <- new.env(parent = emptyenv())

observeEvent(input$file, {
  df <- read.csv(input$file$datapath)
  data_env$uploaded <- df
  register_protected_data(df)  # runtime registration; no advance column list needed
})
```

A complete runnable host-style Shiny example is installed at:

```{r}
system.file("examples/data_shield_upload_app.R", package = "codeagent")
```

From a development checkout:

```r
devtools::load_all(".")
source("inst/examples/data_shield_upload_app.R")
```

It demonstrates three outcomes after upload: bulk rows withheld by `row_cap`,
a single indexed value withheld by `value_match`, and a harmless shape summary
passed through.

> **Current P0.5 limitation:** the protected-value index is process-global.
> The upload demo is therefore suitable for a single-user/development process.
> A shared-process multi-user Shiny deployment requires indexing by session/chat
> before production use (tracked in the Data Shield plans).

Illustrative behaviour of the P0 row-cap (implemented):

| tool output | P0 action |
|---|---|
| `print(mtcars)` (bulk rows) | capped → shape summary |
| a tibble print (`# A tibble: 320 x 12`) | capped |
| `print(nrow(df))` → `320` | passes |
| a status message | passes |
| `print(summary(fit))` | passes (not row data) |

## Roadmap

- **P0.5 — `value_match` (available)**: deterministically catches *targeted*
  leaks the row-cap lets through (e.g. printing one patient's name), by matching
  tool output against high-entropy values registered with
  `register_protected_data()`.
- **P1 — `DescribeData` + `register_protected_data()`**: the model's sanctioned,
  hardened view of data (schema + suppressed/k-anonymised summaries), plus the
  protected-data registry.
- **P2 — reviewer (small model), sandbox (folder + no-network), differential
  privacy for distributions (opt-in).**

## Honest limits

Data Shield **reduces** disclosure risk; it does not eliminate it. Deterministic
detectors (row-cap, `value_match`, regex) miss adversarially obfuscated egress
(e.g. base64-encoding data before printing); those are mitigated — not solved —
by the ingress blacklist and the no-network sandbox. The strongest guarantees
come from the structural layers (metadata-only feeding + no-network execution),
with scanning as defense-in-depth.
