---
title: "Data exploration with codeagent"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data exploration with codeagent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, purl=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Language:** English | [简体中文](data-exploration-cn.html)

## Standard, WEAR, and protected-data paths

codeagent has two general data-exploration experiences and a separate safe
metadata path for protected data:

| | Standard session | WEAR session (`wear_explore()`) | Data Shield `DescribeData` |
|---|---|---|---|
| Entry point | `codeagent_app()` / `codeagent_console()` | `wear_explore(data = ...)` | `codeagent_client(data_shield = ...)` |
| Tool | `ExploreData` is registered by default | Re-registers `ExploreData` for `data`, adds `GenerateReport` | `DescribeData` for registered protected datasets |
| Model behavior | General-purpose agent | WEAR prompt requires a **Next steps** section | Filtered schema/metadata only; no arbitrary query code |
| Typical use | Data questions among coding tasks | Focused, stateful EDA | Discovering protected schema without exposing rows |

**Rule of thumb:** use `wear_explore()` when exploration is the task. Use a
standard session when data questions are one part of broader work. For protected
data, use Data Shield and `DescribeData`; `ExploreData` is not a security
boundary.

## Standard sessions: `ExploreData` is on by default

A normally constructed client registers `ExploreData` unless
`explore_data` is explicitly set to `false` in settings. It reads named
data.frames from `.GlobalEnv` by default. Calling
`register_explore_data_tool(chat, envir = ...)` explicitly remains useful to
install it on a bare chat or bind it to another environment.

```{r, purl=FALSE}
library(codeagent)

client <- codeagent_client(permission_mode = "bypass", btw_groups = NULL)
register_explore_data_tool(client$chat)

codeagent(client, "How many rows are in mtcars and what are the columns?")
codeagent(client, "What is the average mpg per cylinder count?")
```

To opt out globally or per project:

```json
{
  "explore_data": false
}
```

`explore_data_tool(envir = .GlobalEnv)` exposes three model arguments:
`data_name` (required), `question` (optional), and `code` (optional). Without
`code`, it returns `ellmer::df_schema(df)`. With `code`, it parses and evaluates
the expression and returns a table artifact for a data.frame or text for other
results.

The tool is annotated as read-only and is currently classified as a read tool by
the central permission gate, so it is allowed without prompting unless a host
adds a tool override. That classification describes intended use; it is not an
R sandbox guarantee. See [Security and execution limits](#security-and-execution-limits).

## WEAR sessions: dedicated exploration

`wear_explore(data = NULL, client = NULL, mode = c("repl", "shiny"), ...)`
starts a Write/Execute/Analyze/Regroup session. The default mode is `"repl"`.
When `client` is `NULL`, it creates one with `permission_mode = "bypass"`.
`data = NULL` uses `.GlobalEnv`; an environment is used directly; another
list-like value is converted with `list2env(as.list(data), parent = .GlobalEnv)`.
Use a named list so every data.frame has a stable tool-visible name.

The function then:

1. registers `ExploreData` against that environment;
2. registers the model-callable `GenerateReport` tool;
3. appends the WEAR system prompt if it is not already present; and
4. enters `codeagent_console()` or `codeagent_app()`.

For each user question the prompt instructs the model to:

1. **W — Write:** generate dplyr/base R code;
2. **E — Execute:** call `ExploreData`;
3. **A — Analyze:** interpret patterns, outliers, and unexpected findings; and
4. **R — Regroup:** end every response with exactly three numbered suggestions
   under **Next steps**.

```
wear_explore(data = ...)
  - resolve data into an environment
  - register ExploreData for that environment
  - register GenerateReport
  - append the WEAR system prompt
  - enter codeagent_console() or codeagent_app()

Each turn is prompted to follow W-E-A-R:
  W  Write     model writes dplyr/base R code
  E  Execute   ExploreData evaluates code in a new child environment
                 - code supplied -> table artifact or printed value
                 - no code       -> ellmer::df_schema(df)
  A  Analyze   model interprets the result
  R  Regroup   model ends with "Next steps" and exactly 3 suggestions
       |
       v  repeat until the user stops
  /report -> model is instructed to call GenerateReport
```

```{r, purl=FALSE}
# CLI session (blocks until you type /exit or Ctrl+C)
wear_explore(data = list(mtcars = mtcars))

# Shiny UI
wear_explore(
  data = list(sales = sales_df, products = products_df),
  mode = "shiny"
)
```

`wear_explore()` invisibly returns the client after the interactive frontend
ends.

### Exporting to Quarto

In a WEAR session, `/report` is locally recognized as the built-in `report`
skill. The REPL/Shiny skill path expands that skill into a model prompt, which
instructs the model to call the registered `GenerateReport` tool; it does not
invoke the exporter directly in local command dispatch. Asking to save or
export the analysis should invoke the same tool. You can also call
`generate_wear_report()` directly.

```{r, purl=FALSE}
# Capture the client to generate the report after the session ends
client <- wear_explore(data = list(mtcars = mtcars))

path <- generate_wear_report(
  client,
  path  = "mtcars-analysis.qmd",
  title = "Motor Trend Car Analysis"
)

# Render with the quarto CLI
system(paste("quarto render", path))
```

`generate_wear_report(client, path = ..., title = "Data Exploration Report")`
reads the chat turns and writes YAML with `code-fold: true`, `toc: true`, and
`execute: {echo: true, eval: false}`. User text becomes `##` headings and
assistant text becomes prose. Tool-result text that does not begin with `[` is
written as a blockquote.

The exporter writes an R code chunk only when a `ContentToolResult` contains
`extra$code`. The current `ExploreData` implementation returns typed artifacts
but does **not** attach `extra$code`, so ordinary WEAR sessions do not currently
preserve the generated query code in the `.qmd`. Consequently, treat the export
as a conversation report, not a fully reproducible analysis script, unless the
recorded tool results were produced by a tool that supplies `extra$code`.

## `DescribeData` for protected datasets

`DescribeData` is distinct from `ExploreData`. It is installed by an active Data
Shield when the describe strategy is enabled, and it only describes data.frames
registered on that shield. It accepts an optional `data_name`; omission is
allowed when exactly one protected dataset is registered.

```{r, purl=FALSE}
shield <- DataShield$new(strategies = list(
  shield_describe(k_anon = 5),
  shield_egress(max_rows = 0),
  shield_regex(on_fail = "redact")
))
shield$register_data(mtcars, name = "cars")

client <- codeagent_client(data_shield = shield)
# The model can call DescribeData(data_name = "cars").
```

With `shield_describe()` defaults (`distributions = "off"`, `k_anon = 5`,
`category_max = 20`, `category_ratio = 0.2`), `DescribeData` returns filtered
schema information, sensitivity classes, missingness presence, safe numeric/date
ranges, and sufficiently supported low-cardinality labels. It returns no rows,
free-text examples, distributions, or category counts. Identifier and
quasi-identifier values are suppressed and rare labels become
`<rare suppressed>`.

`distributions = "on"` explicitly enables real per-category counts.
`distributions = "dp"` emits Laplace-noised category counts and consumes the
per-dataset budget (`dp_epsilon = 1`, `dp_budget = 5` by default), then silently
degrades to labels-only output after exhaustion. Numeric/date summaries are
unchanged by these distribution modes.

When a shield is supplied to `codeagent_client()`, filtered schemas for datasets
registered before client construction are also placed in the system prompt.
Building that schema block uses the same DP describe path: in `"dp"` mode it
consumes and persists `dp_epsilon` for each eligible categorical column.
Consequently, initial prompt construction and repeated
`refresh_data_shield_context(client)` calls can exhaust the dataset budget.
After registering runtime uploads, call `refresh_data_shield_context(client)` to
refresh that prompt context. `DescribeData` remains the live authoritative
fallback.

Do not register protected data only with `ExploreData`: it can execute arbitrary
R and can return row-level values. Data Shield wraps tool egress and adds the
protected-data metadata path, but the shield must be configured and the dataset
must be registered before relying on those controls.

## Environment context injection

This independent setting adds `.GlobalEnv` object names and data.frame schemas
to the first system reminder so the model knows what exists without first
calling a tool:

```json
// ~/.codeagent/settings.json
{
  "inject_r_env": true
}
```

It does not register data with Data Shield and does not change `ExploreData`
execution semantics.

## Security and execution limits

`ExploreData` evaluates model-generated R with `eval(parse(...))` in a new
environment whose parent is the supplied data environment. It binds the selected
data.frame in that child. Ordinary replacement such as `df$x <- ...` uses the
child binding/copy-on-modify and is tested not to change the source data.frame.

However, this is namespace isolation, **not a security sandbox**. The evaluated
code can resolve functions through parent environments and can potentially use
explicit parent assignment, mutable/reference objects, filesystem functions,
process execution, or network functions available in the R process. The
`read_only_hint = TRUE` and `open_world_hint = FALSE` annotations are metadata,
not enforcement. Use only with a trusted model and environment, add an explicit
permission override if required by the host, and use Data Shield/
`DescribeData` rather than `ExploreData` for protected data. The general `RunR`
tool remains separately classified as execution-capable and goes through its
own permission decision.
