---
title: "Data exploration with codeagent"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data exploration with codeagent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

codeagent includes a dedicated data exploration mode inspired by
[querychat](https://github.com/posit-dev/querychat). The agent can answer
natural-language questions about data.frames in your R session without
writing raw SQL or dplyr code yourself.

## Quick start

```{r}
library(codeagent)

# Register the ExploreData tool on a client
client <- codeagent_client(
  permission_mode = "bypass",
  btw_groups = NULL  # disable btw for a focused data session
)
register_explore_data_tool(client$chat)

# Now ask questions in the chat
codeagent(client, "How many rows are in mtcars and what are the columns?")
codeagent(client, "What is the average mpg for each number of cylinders?")
codeagent(client, "Which cars have mpg > 25? Show name and mpg.")
```

## Using /explore in the Shiny app

Type `/explore mtcars` in the chat to enter data exploration mode:

```
/explore mtcars What is the relationship between weight and mpg?
```

The agent will:
1. Get the schema of `mtcars`
2. Write dplyr code to answer your question
3. Execute it in a sandboxed environment (read-only, never modifies data)
4. Display results as an interactive table

## Environment context injection

Enable automatic schema injection so the agent always knows what
data.frames are in your session:

```json
// ~/.codeagent/settings.json
{
  "inject_r_env": true
}
```

With this setting, the system-reminder includes a summary of all objects
in `.GlobalEnv` on the first turn -- no tool call needed to discover them.

## Security

The `ExploreData` tool:
- Executes code in a **sub-environment** of `.GlobalEnv` (cannot modify globals)
- Is marked `read_only_hint = TRUE` (auto-allowed in `default` permission mode)
- Does not have access to the file system or network

For full RunR capabilities (file I/O, network), use the standard RunR tool.
