---
title: "Data Shield: strict data-safety mode (design preview)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Shield: strict data-safety mode (design preview)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status — P0/P0.5/P1/P1.5/C2/C5-policy available; fuller design in
> progress.** The egress row-cap, protected-value matching, strict DescribeData,
> ordered scanners, universal pre-tool ingress, per-tool policy, and portable
> path/symlink sandbox are wired. Reviewer, full OS execution adapter, and
> `distributions="on"/"dp"` remain roadmap. Off by default
> (`data_shield = NULL`).

## Why

When codeagent is embedded as a backend over **sensitive data** (clinical,
financial, PII), the guarantee we want is: the LLM may see **metadata and
descriptive summaries**, but **never raw row-level data**. Data Shield is an
**opt-in, pluggable valve** that enforces this — off by default
(`data_shield = NULL`), composed from independent strategies when on.

## The core: two edges

Strip away the agent machinery and data can reach the LLM through **only two
inbound edges**. Guard both and you have guarded everything:

1. **Prompt / system-prompt content** — including the framework's own
   **ambient-context auto-injection** (codeagent can inject a summary of R
   session objects into the system reminder). This is the part we control and
   must keep **schema-only** (names / types / dimensions), never values.
2. **Tool results** — whatever a tool returns and that is fed back to the model.

Everything else reduces to these two (RAG and errors arrive as one of them;
model-generated content is outbound, not inbound). The guarantee applies
**recursively to sub-agents** — each has its own two edges.

> Blind spot to handle separately: an **image/multimodal** tool result (a
> rendered table/plot of raw rows) bypasses text scanning.

## Design: composable strategies (not a fixed mode)

`data_shield` accepts `NULL` (off), an ordered strategy list (creates one
private `DataShield` R6), or an explicit `DataShield` instance (for upload and
intentional session/thread sharing). Implemented strategies are shown first;
planned strategies are listed separately so the examples never imply they
already exist.

```{r}
# Implemented now:
client <- codeagent_client(..., data_shield = list(
  shield_describe(k_anon = 5),
  shield_egress(detectors = c("row_cap", "value_match"),
                max_rows = 0, on_fail = "block"),
  shield_regex(on_fail = "redact"),
  shield_ingress(langs = c("r", "python", "bash"), on_fail = "ask"),
  shield_tool_policy(rules = list(
    KMPlot = list(ingress = "scan", egress = "bypass"),
    DangerousExport = list(execution = "deny")
  )),
  shield_sandbox(project_root = getwd(), backend = "policy")
))

# Roadmap (not implemented yet):
# shield_reviewer(...), shield_narrow_tools()
```

The main boundary is **edge 2 (tool results)**; sandbox, ingress blacklist,
reviewer, and tool narrowing are defense-in-depth, not the boundary.

## Current parameter reference

### `data_shield` input

| Value | Effect |
|---|---|
| `NULL` | Shield completely off; existing codeagent behaviour unchanged |
| `list(shield_*())` | Strategies run in list order; codeagent creates one private `DataShield` R6 for this client |
| `DataShield$new(...)` | Explicit lifecycle for uploads and deliberate sharing among selected chats |

### `shield_egress()` — core tool-result boundary

| Parameter | Default | Actual effect |
|---|---|---|
| `detectors` | `c("row_cap", "value_match")` | `row_cap` catches bulk tabular output; `value_match` catches indexed high-entropy values |
| `max_rows` | `0` | On a bulk/tabular match, retain zero raw printed lines and return only a withheld/blocked notice. `5` would deliberately expose the first five printed lines |
| `on_fail` | `"redact"` | `redact`: withheld notice; `block`: blocked notice. Neither asks a user yet |

`max_rows = 0` does **not** block `print()` generally. `print(nrow(df))`, status
messages, model summaries, plots, and errors pass unless another detector finds
sensitive content. It triggers only when output has a data.frame/tibble or
many-line rectangular-table shape.

### `shield_describe()` — strict model-safe metadata

| Parameter | Default | Actual effect |
|---|---|---|
| `distributions` | `"off"` | Implemented strict mode: no histograms, quantiles, means, category counts, rows, or free-text examples. `on`/`dp` are roadmap and fail explicitly |
| `k_anon` | `5` | Category labels supported by fewer than k rows become `<rare suppressed>` |
| `category_max` | `20` | Maximum distinct character values for categorical treatment |
| `category_ratio` | `0.2` | Maximum distinct/non-missing ratio for character-categorical treatment; otherwise `free_text` |

Sensitivity still clamps output: `identifier`/`quasi` values stay suppressed;
`measure`/`open` may show numeric/date min–max and safe category labels without
counts.

### `shield_regex()` — unregistered PII/secrets

| Parameter | Default | Actual effect |
|---|---|---|
| `patterns` | `NULL` | Optional named regular-expression rules using PCRE (Perl-Compatible Regular Expression) syntax, e.g. `c(study_id = "STUDY-[0-9]+")`; appended to defaults when enabled |
| `include_defaults` | `TRUE` | Email, phone-like, common-token-prefix, and 18-character identity-number rules |
| `replacement` | `"[REDACTED]"` | Marker inserted once per merged matching span |
| `on_fail` | `"redact"` | `redact`: preserve safe surrounding text; `block`: replace the whole result |
| `ignore_case` | `TRUE` | Case-insensitive matching for all rules |


### `shield_ingress()` — scan every tool call before execution

| Parameter | Default | Actual effect |
|---|---|---|
| `langs` | `c("r", "python", "bash")` | Select built-in rules for each code/shell language |
| `patterns` | `NULL` | Optional named custom regex rules appended to defaults |
| `include_defaults` | `TRUE` | Include serialization/base64/network/data-file-display and protected-name preview rules |
| `on_fail` | `"block"` | `block`: reject the tool call; `ask`: force the existing permission approval UI/callback |
| `ignore_case` | `TRUE` | Case-insensitive matching |

Ingress scans **all** tool arguments before the usual read/write/exec capability
fast paths, including unknown and read-only tools. It does not ban ordinary
reading: defaults focus on high-confidence read-and-display, serialization,
encoding, and network-transfer patterns. It is a cheap pre-filter; egress
scanning remains the primary boundary because code can be obfuscated.

### `shield_tool_policy()` — exact/glob trust and deny rules

| Setting | Meaning |
|---|---|
| `default="scan"` | Every tool is scanned unless a rule overrides it |
| `execution="deny"` | Reject the tool before execution |
| `ingress="bypass"` | Skip Shield argument scanning, but still apply permission gate |
| `egress="bypass"` | Return that tool's output without Shield filtering; audit every bypass |
| `egress="deny"` | Replace the result with an explicit policy-denied notice |

Rules support exact names and `*` globs. Exact wins; otherwise the first matching
glob wins. For example, `KMPlot` may bypass egress when its developer guarantees
all outputs are LLM-safe, while `btw_tool_docs_*` can receive a broader trusted
rule. This policy never bypasses codeagent's independent permission system.

### `shield_sandbox()` — portable containment without crippling the agent

| Parameter | Default | Actual effect |
|---|---|---|
| `project_root` | `getwd()` | Project root allowed by path policy |
| `protected_paths` | none | Extra registered data roots; longest matching root controls mode |
| `temp_root` | new session temp | Isolated temporary root |
| `modes` | project `rwx`, data `rw`, temp `rwx` | Logical Shield capabilities (not chmod bits) |
| `process_exec` | `TRUE` | Preserve RunR/Bash/Python; FALSE blocks exec tools |
| `network` | `"tool_policy"` | Let tool policy decide; `"deny"` blocks net capability |
| `symlink_escape` | `"deny"` | Resolve real paths and reject links escaping allowed roots |
| `backend` | `"auto"` | `policy`, `auto`, or `required` |
| `on_unavailable` | `"policy"` | Full OS adapter unavailable → policy fallback; `block` fails closed for exec/net |

Current implementation is a portable central-gate path/capability policy. It is
not advertised as kernel isolation: the capability probe found user/network/
mount namespaces but no bubblewrap/container, and plain `unshare` still sees the
host filesystem. A future full adapter must move exec tools out of process.
### `DataShield$new()` direct lifecycle

Direct constructor parameters (`max_rows`, `distributions`, `k_anon`,
`category_max`, `category_ratio`, `audit_max`) create the default DescribeData +
core egress configuration when `strategies = NULL`. `audit_max` defaults to 1000
non-sensitive decision events; `0` disables recording. Supplying
`strategies = list(...)` enables **only** listed strategies and preserves list
order. Use `shield$register_data()`, `$install()`, `$describe()`, `$audit()`,
`$clear_audit()`, `$clear()`, and `$close()` for dynamic/session-owned workflows.

## Plain-language glossary

| Term | Plain meaning |
|---|---|
| **ingress** | A tool call's name/arguments entering local execution; scanned before the tool runs |
| **egress** | Content leaving a local tool and about to enter the LLM |
| **row-cap** | A limit on how many printed table lines may pass; `0` means no raw line |
| **value-match** | Exact matching against high-entropy values indexed from registered protected data |
| **PII** | Personally identifiable information, such as email, phone, identity number, or name |
| **kind** | What a registered asset is (dataset/spec/document/synthetic), not its R data type or access level |
| **provenance** | A verifiable source tag showing which registered asset produced a result |
| **raw access** | Content may enter an LLM edge without row/value restrictions; optional secret/PII scanning may still apply |
| **regex** | Regular expression: a text pattern such as `STUDY-[0-9]+` |
| **PCRE** | Perl-Compatible Regular Expression, the regex syntax used by R with `perl=TRUE` |
| **span** | Start/end character positions of a detected sensitive substring, enabling precise replacement |
| **k-anonymity threshold** | Do not expose a category label unless at least `k` rows support it |
| **fail closed** | If a safety scanner fails, block output instead of allowing it |
| **R6** | R's mutable object system; one `DataShield` owns private datasets/index/lifecycle |

## Non-sensitive audit log

`shield$audit()` returns an in-memory data.frame of policy decisions:

| Field | Meaning |
|---|---|
| `timestamp` | UTC event time |
| `edge` | `ingress` (before tool) or `egress` (before LLM) |
| `tool_name`, `tool_call_id` | Non-sensitive correlation identifiers |
| `strategy` | `row_cap`, `value_match`, `regex`, `ingress`, or custom scanner name |
| `action`, `reason` | `redact`/`block`/`ask` and a rule/reason label |
| `match_count`, `score` | Number of matches and normalized risk score |

It **never stores** raw tool arguments/results, matched values, data rows, span
text, or hashes. `audit_max` bounds memory (oldest events are dropped); use
`shield$clear_audit()` to clear it. Each `DataShield` R6 has its own log, so
session/thread isolation matches the policy instance.

```r
recent <- shield$audit(limit=100)
shield$clear_audit()
```

## Data Asset Policy: what it is × what the LLM may see

Asset content type and LLM access are orthogonal:

| `kind` | Default prompt | Default egress | Typical use |
|---|---|---|---|
| `dataset` | `schema` | `scan` | patient/analysis data |
| `spec` | `raw` | `scan` | ADaM spec, SDTMIG, public dictionaries |
| `synthetic` | `raw` | `scan` | dummy/edge-case test data |
| `document` | `scan` | `scan` | ordinary reference documents |

| Access | Meaning |
|---|---|
| `none` | content unavailable to that LLM edge |
| `schema` | only strict DescribeData-style metadata |
| `scan` | content must pass configured Shield scanners |
| `raw` | bypass row/value restrictions; baseline secret/PII regex still applies unless explicitly disabled |

```r
shield$register_asset(
  adam_spec,
  name = "adam_spec",
  kind = "spec",
  llm_access = list(prompt = "raw", egress = "scan"),
  scan_secrets = TRUE,
  reason = "Validated public specification",
  expires = "session"
)

prompt_text <- shield$prompt_content("adam_spec")
```

Raw egress never follows from `kind` alone. It requires explicit policy and
provenance:

```r
shield$register_asset(
  adam_spec, name = "adam_spec", kind = "spec",
  llm_access = list(prompt = "raw", egress = "raw"),
  reason = "Validated public specification")

tool_result <- shield$trusted_result(value, source = "adam_spec")
```

An untagged or mixed tool result remains scanned. Raw asset policies require a
reason, expire with the owning DataShield session by default, may set a POSIXct
expiry, and emit bypass audit events. Synthetic raw always keeps baseline
PII/secret scanning; spec raw may explicitly set `scan_secrets = FALSE`.

## P0 — the foundation (available now)

The minimal, deterministic slice that already gives real protection:

```{r}
# Easy entry: strategy specs create one private DataShield R6 for this client.
client <- codeagent_client(chat, data_shield = list(
  shield_describe(k_anon = 5),

  shield_egress(max_rows = 0)
))

# Harness-only client: attach tools, then install its R6 engine.
client <- codeagent_client(chat, register_tools = FALSE,
  data_shield = list(shield_describe(), shield_egress(max_rows = 0)))
chat$register_tool(my_tool)
client$data_shield$install(client$chat)
```

- **Edge 2 — shape-based egress row-cap.** codeagent does **not** inspect code
  or block `print`. It looks only at the **shape** of a tool's returned text:
  output with a data.frame / tibble print signature or a many-row rectangular
  table is truncated to a shape summary; scalars, messages, **model summaries**,
  plots and errors pass through untouched. Content-agnostic and tunable via
  `max_rows`.
- **Edge 1 — ambient stays schema-only.** codeagent's ambient injection already
  emits only `name [data.frame N x M: col:type, ...]` (no values); Data Shield
  keeps it that way.

### Runtime uploads in Shiny

The dataset does **not** need to be known when the app starts. Register it in
the upload observer immediately after reading it; tools may already be attached
and wrapped, because value matching reads the live index at invocation time.

```{r}
# Inside each Shiny server session: one R6 may be shared by selected chats.
shield <- DataShield$new(
  strategies = list(shield_describe(), shield_egress(max_rows = 0)))
data_env <- new.env(parent = emptyenv())

client_factory <- function() {
  codeagent_client(make_chat(), data_shield = shield)
}

observeEvent(input$file, {
  df <- read.csv(input$file$datapath)
  data_env$uploaded <- df
  shield$register_data(df, name = "uploaded") # no advance columns needed
})
```

A complete runnable host-style Shiny example is installed at:

```{r}
system.file("examples/data_shield_upload_app.R", package = "codeagent")
```

From a development checkout:

```r
devtools::load_all(".")
source("inst/examples/data_shield_upload_app.R")
```

It demonstrates five outcomes after upload: bulk rows withheld by `row_cap`, a
single indexed value withheld by `value_match`, unregistered PII redacted by
`shield_regex`, a harmless shape summary passed through, and strict
`DescribeData` metadata with raw identifiers suppressed.

> **Multi-user isolation:** create `DataShield$new()` inside the Shiny server
> function, share it only among the intended chat threads, and register data via
> `shield$register_data()`. Other browser sessions receive separate R6 instances
> and cannot see or influence its index.

Illustrative behaviour of the P0 row-cap (implemented):

| tool output | P0 action |
|---|---|
| `print(mtcars)` (bulk rows) | capped → shape summary |
| a tibble print (`# A tibble: 320 x 12`) | capped |
| `print(nrow(df))` → `320` | passes |
| a status message | passes |
| `print(summary(fit))` | passes (not row data) |


## P1 `DescribeData`: strict safe metadata contract

`DescribeData` is the sanctioned way for the model to understand protected data
without receiving rows. Its output is determined by three orthogonal dimensions:

| Dimension | Values | Purpose |
|---|---|---|
| Global policy | `distributions = "off" / "on" / "dp"` | strict defaults to no distributions; `on` is explicit opt-in; `dp` adds noise + budget |
| Column sensitivity | `identifier / quasi / measure / open` | business role that clamps the maximum disclosure for that column |
| Data type | numeric / factor / character / Date / … | determines the safe representation: range, labels, or free-text marker |

Strict (`distributions = "off"`) matrix:

| Metadata | identifier / quasi | measure / open |
|---|---|---|
| column name, type, missing presence | shown | shown |
| numeric/date min–max | hidden | shown |
| low-cardinality categorical labels | hidden | shown without counts; levels with support `< k` are suppressed |
| real free-text examples | hidden | hidden |
| histogram, quantiles, mean/SD, category counts | hidden | hidden (`on`/`dp` opt-in only) |

A factor is not automatically safe: a factorised subject ID remains an
`identifier`. Character columns receive categorical labels only when they are
low-cardinality, low-uniqueness, non-PII, and every exposed level satisfies the
k-anonymity threshold. Free text never receives real examples in strict mode.

## P1.5 ordered egress scanners

Strategy-list order is the execution order. `shield_regex()` catches sensitive
content even when no data.frame was registered:

```r
client <- codeagent_client(chat, data_shield=list(
  shield_egress(max_rows=0),
  shield_regex(on_fail="redact"),
  shield_regex(patterns=c(study_id="STUDY-[0-9]+"),
               include_defaults=FALSE, on_fail="block")
))
```

Built-ins cover email, phone-like strings, common API-token prefixes and
18-character identity-number shapes. `redact` replaces only matched spans;
`block` discards the whole model-facing result. Custom scanner functions may be
appended with `shield$add_scanner(name, fn)`; invalid scanner results/errors fail
closed.


## C2 universal ingress scanning

`shield_ingress()` is installed into codeagent's existing single central
permission gate; it does not create a competing callback/gate. It scans every
tool's arguments before the read-only fast path. A `block` result raises
`tool_reject`; an `ask` result reuses the current CLI/Shiny approval callback,
including tool-call id correlation.

```r
client <- codeagent_client(chat, data_shield=list(
  shield_ingress(on_fail="ask"),
  shield_egress(max_rows=0),
  shield_regex()
))
```

Defaults intentionally do not reject every `Read` or `print`: `nrow(study)` and
`print("done")` pass, while `head(study)`, `dput(study)`, base64/pickle/JSON
serialization, upload-style curl/requests calls, and shell display of data files
are reviewed or blocked.

## C5 portable sandbox and btw boundary

`shield_sandbox()` deliberately preserves coding capability: project and
session-temp default to `rwx`, protected data defaults to `rw` but may be `rwx`,
and process execution stays enabled. Its current portable backend blocks
explicit paths outside allowed roots, rejects symlink escape, and applies
network/process capability policy in the central gate.

btw is not assumed to provide OS isolation. Its file tools enforce cwd with
`fs::path_has_parent()` but a project-internal symlink to an external file passed
our probe; its RunR executes through `evaluate` in the global environment and
only restores cwd/options/envvars. Data Shield therefore applies uniformly to
native, btw, MCP and host tools.
## Roadmap

- **P0.5 — `value_match` (available)**: deterministically catches *targeted*
  leaks the row-cap lets through (e.g. printing one patient's name), by matching
  tool output against high-entropy values registered with
  `shield$register_data()`.
- **P1 — `DescribeData` + protected-data registry (strict available)**: the
  model's sanctioned hardened view (schema, sensitivity, missing presence,
  measure/open ranges and k-supported labels; no distributions/counts/examples).
  `distributions="on"/"dp"` remain later phases.

## Sub-agent boundary

Foreground sub-agents (`Agent`) inherit the exact same `DataShield` R6 before
any of their tools can return content to the child model. This applies to both
synchronous and concurrent async Agent calls. While a shield is active,
codeagent deliberately skips btw/custom-agent delegation paths that cannot
accept the policy engine.

`BackgroundAgent` and `/bg` currently **fail closed** under Data Shield: their
mirai worker is a separate R process and cannot safely share the session's R6
state or protected-value index. Use foreground `Agent` until a per-owner worker
reconstruction protocol is implemented.
- **P1.5 — ordered scanner pipeline + `shield_regex()` (available)**:
  unregistered PII/secrets are redacted/blocked with precise spans; custom
  scanner failures fail closed.
- **C2 — `shield_ingress()` (available)**: all tool arguments pass the central
  permission gate; deterministic high-confidence rules block or force approval.
- **C5 — portable `shield_sandbox()` (available)**: project/temp `rwx`, protected
  data `rw` by default, realpath/symlink containment and policy fallback; full OS
  process adapter remains roadmap.
- **P2 — reviewer (small model), sandbox (folder + no-network), differential
  privacy for distributions (opt-in).**

## Honest limits

Data Shield **reduces** disclosure risk; it does not eliminate it. Deterministic
detectors (row-cap, `value_match`, regex) miss adversarially obfuscated egress
(e.g. base64-encoding data before printing); those are mitigated — not solved —
by the ingress blacklist and the no-network sandbox. The strongest guarantees
come from the structural layers (metadata-only feeding + no-network execution),
with scanning as defense-in-depth.
