---
title: "Data Shield: strict data-safety mode (design preview)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Shield: strict data-safety mode (design preview)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

**Language:** English | [简体中文](data-shield-cn.html)

> **Status — P0/P0.5/P1/P1.5/C2/C5-policy available; fuller design in
> progress.** The egress row-cap, protected-value matching, strict DescribeData,
> ordered scanners, universal pre-tool ingress, per-tool policy, promise-backed
> egress approval, portable path/symlink sandbox, optional sanitized-code
> reviewer, and DP-noised categorical counts (`distributions="dp"`) are wired.
> Full OS execution adapter and differential privacy for numeric/continuous
> statistics remain roadmap. Off by default (`data_shield = NULL`).

## Why

When codeagent is embedded as a backend over **sensitive data** (clinical,
financial, PII), the guarantee we want is: the LLM may see **metadata and
descriptive summaries**, but **never raw row-level data**. Data Shield is an
**opt-in, pluggable valve** that enforces this — off by default
(`data_shield = NULL`), composed from independent strategies when on.

## The core: three edges

Strip away the agent machinery and data crosses the model boundary at **three
edges**. Data Shield installs a gate at each one:

1. **User/model input** (edge 1, input gate) — typed text and text-bearing
   attachments are scanned before the model sees them. Framework-owned protected
   dataset context is separately generated as filtered schema-only metadata
   (names, types, dimensions, and policy-safe summaries), never raw rows.
2. **Tool traffic** (edge 2, tool gate) — tool arguments are checked before
   execution and tool results are filtered before they are fed back to the model.
3. **The model's final reply** (edge 3, output gate) — the complete reply is
   scanned before it reaches the user, since the model may reproduce a protected
   value inferred from an edge-2 aggregate even when the user's input was clean.

RAG content, attachments, and errors must enter through one of these edges. The
same policy applies recursively to owned foreground sub-agents.

> **Image boundary:** image prompt attachments are not scanned by default. An
> opt-in OCR hook (`data_shield_ocr_scanner()`) can scan text baked into those
> images at edge 1. Image/multimodal **tool results** still bypass text egress
> scanning and require a host-provided image-result control.

### Architecture at a glance

```
 Host/user registers protected data (local only)
     │ shield$register_data(df, name, sensitivity)
     ├────────► filtered schema block / DescribeData ───────────────┐
     └────────► local high-entropy value index                      │
                                                                   ▼
 user text + text attachments ─► [edge 1 input gate] ───────────► model
                                                                   │
 tool call ─► [edge 2 ingress: policy → patterns → reviewer] ─► execute
                                                                   │
 model ◄── [edge 2 egress: row_cap → value_match → scanners] ◄── result
   │
   └──────── final reply ─► [edge 3 output gate] ─► user/browser

 Sandbox and tool narrowing are defense-in-depth. The permission gate remains
 independent, and a Shield bypass never bypasses permission checks.
```

### Setup used by every snippet below

Every code block in this vignette continues from a `chat` (a bare **ellmer**
`Chat`, not the `CodeagentClient` wrapper) and a `shield` (a `DataShield`
instance) built once, up front:

```r
chat <- ellmer::chat_openai_compatible(
  base_url    = Sys.getenv("CODEAGENT_BASE_URL"),
  model       = Sys.getenv("CODEAGENT_MODEL"),
  credentials = function() Sys.getenv("CODEAGENT_API_KEY"))
shield <- DataShield$new()
```

`codeagent_client(chat, ...)` does **not** clone `chat` — `client$chat` is the
exact same object you passed in (mutable R6/environment semantics), so a
snippet may call either `chat$register_tool(...)` or
`client$chat$register_tool(...)` interchangeably; this vignette uses the bare
`chat` variable throughout for brevity.

## Design: composable strategies (not a fixed mode)

`data_shield` accepts `NULL` (off), an ordered strategy list (creates one
private `DataShield` R6), or an explicit `DataShield` instance (for upload and
intentional session/thread sharing). Implemented strategies are shown first;
planned strategies are listed separately so the examples never imply they
already exist.

```{r}
# Implemented now:
client <- codeagent_client(chat, data_shield = list(
  shield_describe(k_anon = 5),
  shield_egress(detectors = c("row_cap", "value_match"),
                max_rows = 0, on_fail = "block"),
  shield_regex(on_fail = "redact"),
  shield_ingress(langs = c("r", "python", "bash"), on_fail = "ask"),
  shield_tool_policy(rules = list(
    KMPlot = list(ingress = "scan", egress = "bypass"),
    DangerousExport = list(execution = "deny")
  )),
  shield_sandbox(project_root = getwd(), backend = "policy"),
  shield_reviewer(model = Sys.getenv("CODEAGENT_FAST_MODEL"),
                  scope = c("exec", "write", "net"), on_risk = "ask")
))

# Roadmap (not implemented yet):
# shield_narrow_tools()
```

The main boundary is **edge 2 (tool results)**; sandbox, ingress blacklist,
reviewer, and tool narrowing are defense-in-depth, not the boundary.

## Current parameter reference

### `data_shield` input

| Value | Effect |
|---|---|
| `NULL` | Shield completely off; existing codeagent behaviour unchanged |
| `list(shield_*())` | codeagent creates one private `DataShield` R6; egress scanners follow list order, while ingress/reviewer/sandbox/tool-policy strategies run at their own phases |
| `DataShield$new(...)` | Explicit lifecycle for uploads and deliberate sharing among selected chats |

### Input/output gates and scanner defaults

The input and final-response gates are client settings rather than `shield_*()`
strategies. Their secure defaults are:

| Setting | Default | Effect |
|---|---|---|
| `data_shield_prompt_on_fail` | `"redact"` | Redact matched spans in user text; `"block"` rejects the turn |
| `data_shield_input_scanners` | `c("value_match", "regex")` | Scan registered protected values and PII/token shapes |
| `data_shield_response_on_fail` | `"redact"` | Redact the finalized model reply; `"block"` replaces it with a blocked notice |
| `data_shield_output_scanners` | `c("value_match", "regex")` | Apply the same two detectors to the final reply |
| `data_shield_image_scanner` | `NULL` | Images are not scanned unless the host supplies a hook |

Unknown scanner names and scanner exceptions fail closed. Input `"ask"` has no
separate approval path and degrades to redaction. Text-bearing attachments are
scanned; if their text cannot be extracted, the turn is blocked because an
immutable attachment cannot be safely rewritten. When a shield is active, CLI
and Shiny streaming buffer the complete model reply, scan it, and only then emit
the safe text once—there is no plaintext-first stream.

To opt into OCR for prompt images:

```r
client$settings$data_shield_image_scanner <-
  data_shield_ocr_scanner(client$data_shield)
```

`tesseract` is an optional dependency. If it is absent, the ready-made scanner
degrades to pass (the image blind spot remains). Once OCR is available and the
scanner is configured, unreadable images, OCR errors, and OCR-text scan errors
block; detected protected text blocks by default.

### `shield_egress()` — core tool-result boundary

| Parameter | Default | Actual effect |
|---|---|---|
| `detectors` | `c("row_cap", "value_match")` | `row_cap` catches bulk tabular output; `value_match` catches indexed high-entropy values |
| `max_rows` | `0` | On a bulk/tabular match, retain zero raw printed lines and return only a withheld/blocked notice. `5` would deliberately expose the first five printed lines |
| `on_fail` | `"redact"` | `redact`: withheld notice; `block`: blocked notice; `ask`: pause before LLM delivery |
| `allow_raw_approval` | `FALSE` | When asking, show only Redact/Block; TRUE adds dangerous Raw once |
| `approval_timeout` | `60` | Async seconds before automatic redact |

`max_rows = 0` does **not** block `print()` generally. `print(nrow(df))`, status
messages, model summaries, plots, and errors pass unless another detector finds
sensitive content. It triggers only when output has a data.frame/tibble or
many-line rectangular-table shape.

With `on_fail="ask"`, the raw result remains local while the callback/UI receives
only tool name/id, strategy, reason label, match count, score, timeout and whether
raw-once is enabled. Redact is the safe default for no callback, errors, invalid
choices, ESC and timeout. Raw-once applies to one result only and is audited.

### `shield_describe()` — strict model-safe metadata

| Parameter | Default | Actual effect |
|---|---|---|
| `distributions` | `"off"` | `"off"`: no counts, category labels only. `"on"`: real per-category counts (no privacy protection). `"dp"`: Laplace-noised per-category counts, spending `dp_epsilon` from a per-dataset `dp_budget`; degrades to `"off"`-style labels once the budget is exhausted. **Numeric/date/logical/free-text columns are unchanged in all three modes** — see below |
| `k_anon` | `5` | Category labels supported by fewer than k rows become `<rare suppressed>` (applies in all three `distributions` modes, before any count is added) |
| `category_max` | `20` | Maximum distinct character values for categorical treatment |
| `category_ratio` | `0.2` | Maximum distinct/non-missing ratio for character-categorical treatment; otherwise `free_text` |
| `dp_epsilon` | `1` | Privacy cost charged per `describe()`/schema-block call for each categorical column exposed under `"dp"`; split across that column's surviving categories. Only meaningful with `"dp"` |
| `dp_budget` | `5` | Total per-dataset epsilon budget under `"dp"` — a one-time allowance, no time-window reset. Only meaningful with `"dp"` |

Sensitivity still clamps output: `identifier`/`quasi` values stay suppressed;
`measure`/`open` may show numeric/date min–max and safe category labels, with
counts added under `"on"`/`"dp"`.

**Scope limit (v1)**: only categorical columns get counts. Numeric columns
(mean/sum/quantiles) are not covered — differentially-private release of a
continuous statistic needs a *clipping bound* to calibrate its noise, and that
bound must not be derived from the private data's own min/max (doing so would
leak privacy strength from the private data itself, a well-known DP pitfall).
Until a host can supply a real, data-independent bound per numeric column,
`range=[min, max]` is shown unchanged in all three `distributions` modes.
Logical and free-text columns are likewise unchanged. Track:
`references/plan/31x-dp-distributions.TODO.md`.

Exhausting a dataset's `dp_budget` is silent and permanent for that dataset's
lifetime (until it is re-registered): no error, no raw count, just a quiet
drop back to `"off"`-style labels-only output. Call
`shield$dp_budget_remaining(name)` (or with no argument, a named vector for
every registered dataset) to check remaining budget, e.g. for a host-side
"privacy budget: 2/5 remaining" indicator. Every consumption/exhaustion is
also recorded in the audit log (`strategy = "dp_budget"`).

```r
dp_metadata <- shield_describe(distributions = "dp", dp_epsilon = 1, dp_budget = 5)
shield <- DataShield$new(strategies = list(dp_metadata))
shield$register_data(df, name = "study")
shield$describe("study")            # spends 1 epsilon if "arm" is categorical
shield$dp_budget_remaining("study")  # 4
```

### `shield_regex()` — unregistered PII/secrets

| Parameter | Default | Actual effect |
|---|---|---|
| `patterns` | `NULL` | Optional named regular-expression rules using PCRE (Perl-Compatible Regular Expression) syntax, e.g. `c(study_id = "STUDY-[0-9]+")`; appended to defaults when enabled |
| `include_defaults` | `TRUE` | Email, phone-like, common-token-prefix, and 18-character identity-number rules |
| `replacement` | `"[REDACTED]"` | Marker inserted once per merged matching span |
| `on_fail` | `"redact"` | `redact`: preserve safe surrounding text; `block`: replace the whole result |
| `ignore_case` | `TRUE` | Case-insensitive matching for all rules |


### `shield_ingress()` — scan every tool call before execution

| Parameter | Default | Actual effect |
|---|---|---|
| `langs` | `c("r", "python", "bash")` | Select built-in rules for each code/shell language |
| `patterns` | `NULL` | Named regex rules; a name matching a built-in **replaces** that rule, a new name is **added**. Host file-based blacklists read their own file into a named vector and pass it here |
| `include_defaults` | `TRUE` | Include the built-in per-language rule set (`.DATA_SHIELD_INGRESS_RULES`): serialization/encoding, pandas/R writers, network transfer (incl. `nc`/`scp`/`/dev/tcp`), data-file display, and protected-name preview |
| `on_fail` | `"block"` | `block`: reject the tool call; `ask`: force the existing permission approval UI/callback |
| `ignore_case` | `TRUE` | Case-insensitive matching |

Ingress scans **all** tool arguments before the usual read/write/exec capability
fast paths, including unknown and read-only tools. It does not ban ordinary
reading: defaults focus on high-confidence read-and-display, serialization,
encoding, and network-transfer patterns. It is a cheap pre-filter; egress
scanning remains the primary boundary because code can be obfuscated.

### `shield_tool_policy()` — exact/glob trust and deny rules

| Setting | Meaning |
|---|---|
| `default="scan"` | Every tool is scanned unless a rule overrides it |
| `execution="deny"` | Reject the tool before execution |
| `ingress="bypass"` | Skip Shield argument scanning, but still apply permission gate |
| `egress="bypass"` | Return that tool's output without Shield filtering; audit every bypass |
| `egress="deny"` | Replace the result with an explicit policy-denied notice |

Rules support exact names and `*` globs. Exact wins; otherwise the first matching
glob wins. For example, `KMPlot` may bypass egress when its developer guarantees
all outputs are LLM-safe, while `btw_tool_docs_*` can receive a broader trusted
rule. This policy never bypasses codeagent's independent permission system.

### `shield_sandbox()` — portable containment without crippling the agent

| Parameter | Default | Actual effect |
|---|---|---|
| `project_root` | `getwd()` | Project root allowed by path policy |
| `protected_paths` | none | Extra registered data roots; longest matching root controls mode |
| `temp_root` | new session temp | Isolated temporary root |
| `modes` | project `rwx`, data `rw`, temp `rwx` | Logical Shield capabilities (not chmod bits) |
| `process_exec` | `TRUE` | Preserve RunR/Bash/Python; FALSE blocks exec tools |
| `network` | `"tool_policy"` | Let tool policy decide; `"deny"` blocks net capability |
| `symlink_escape` | `"deny"` | Resolve real paths and reject links escaping allowed roots |
| `backend` | `"auto"` | `policy`, `auto`, or `required` |
| `on_unavailable` | `"policy"` | Full OS adapter unavailable → policy fallback; `block` fails closed for exec/net |

Current implementation is a portable central-gate path/capability policy. It is
not advertised as kernel isolation: the capability probe found user/network/
mount namespaces but no bubblewrap/container, and plain `unshare` still sees the
host filesystem. A future full adapter must move exec tools out of process.

### `shield_reviewer()` — optional sanitized-code semantic rail

| Parameter | Default | Actual effect |
|---|---|---|
| `client_factory` | `NULL` | Optional function returning a fresh independent ellmer Chat |
| `model` | `CODEAGENT_FAST_MODEL` | Reviewer model; missing config follows `on_error`, never silently main model |
| `scope` | exec/write/net | Only these tool capabilities incur review cost |
| `on_risk` | `"ask"` | Risk classification becomes ask or block |
| `on_error` | `"ask"` | Missing model, timeout, request/JSON errors become ask or block; no approval→block |
| `backend` | `"remote_sanitized"` | Remote sees only regex/value-sanitized code; `"local_only"` is also supported but requires an explicit local `client_factory` and fails closed without one |
| `timeout` | `30` | Async review timeout seconds |

The reviewer is not a tool and cannot be skipped by the main model. It has no
tools or history; code is delimited as untrusted data and parsed from fixed JSON
(`risk`, `confidence`, `reason`). Deterministic ingress rules run first, so
already-blocked calls incur no model cost.

### `DataShield$new()` direct lifecycle

Direct constructor parameters (`max_rows`, `distributions`, `k_anon`,
`category_max`, `category_ratio`, `audit_max`, `dp_epsilon`, `dp_budget`) create
the default DescribeData + core egress configuration when `strategies = NULL`. `audit_max` defaults to 1000
non-sensitive decision events; `0` disables recording. Supplying
`strategies = list(...)` enables **only** listed strategies. List order controls
the egress scanner pipeline; ingress, reviewer, sandbox, and tool-policy
strategies run at their distinct phases. Use `shield$register_data()`, `$install()`, `$describe()`, `$audit()`,
`$clear_audit()`, `$clear()`, and `$close()` for dynamic/session-owned workflows.

## Plain-language glossary

| Term | Plain meaning |
|---|---|
| **ingress** | A tool call's name/arguments entering local execution; scanned before the tool runs |
| **egress** | Content leaving a local tool and about to enter the LLM |
| **row-cap** | A limit on how many printed table lines may pass; `0` means no raw line |
| **value-match** | Exact matching against high-entropy values indexed from registered protected data |
| **PII** | Personally identifiable information, such as email, phone, identity number, or name |
| **kind** | What a registered asset is (dataset/spec/document/synthetic), not its R data type or access level |
| **provenance** | A verifiable source tag showing which registered asset produced a result |
| **raw access** | Content may enter an LLM edge without row/value restrictions; optional secret/PII scanning may still apply |
| **regex** | Regular expression: a text pattern such as `STUDY-[0-9]+` |
| **PCRE** | Perl-Compatible Regular Expression, the regex syntax used by R with `perl=TRUE` |
| **span** | Start/end character positions of a detected sensitive substring, enabling precise replacement |
| **k-anonymity threshold** | Do not expose a category label unless at least `k` rows support it |
| **fail closed** | If a safety scanner fails, block output instead of allowing it |
| **semantic reviewer** | A separate small model that classifies what sanitized tool code is trying to do; it never sees raw data remotely |
| **R6** | R's mutable object system; one `DataShield` owns private datasets/index/lifecycle |

## Non-sensitive audit log

`shield$audit()` returns an in-memory data.frame of policy decisions:

| Field | Meaning |
|---|---|
| `timestamp` | UTC event time |
| `edge` | Policy boundary, including `prompt`, `ingress`, `egress`, `response`, or `describe` |
| `tool_name`, `tool_call_id` | Non-sensitive correlation identifiers |
| `strategy` | Includes `row_cap`, `value_match`, `regex`, `ingress`, `dp_budget`, `tool_policy`, `asset_policy`, `sandbox`, or a custom scanner name |
| `action`, `reason` | Policy action such as `redact`, `block`, `ask`, `consume`, `exhausted`, `bypass`, or `deny`, plus a rule/reason label |
| `match_count`, `score` | Number of matches and normalized risk score |

It **never stores** raw tool arguments/results, matched values, data rows, span
text, or hashes. `audit_max` bounds memory (oldest events are dropped); use
`shield$clear_audit()` to clear it. Each `DataShield` R6 has its own log, so
session/thread isolation matches the policy instance.

```r
recent <- shield$audit(limit=100)
shield$clear_audit()
```

## Data Asset Policy: what it is × what the LLM may see

Asset content type and LLM access are orthogonal:

| `kind` | Default prompt | Default egress | Typical use |
|---|---|---|---|
| `dataset` | `schema` | `scan` | patient/analysis data |
| `spec` | `raw` | `scan` | ADaM spec, SDTMIG, public dictionaries |
| `synthetic` | `raw` | `scan` | dummy/edge-case test data |
| `document` | `scan` | `scan` | ordinary reference documents |

| Access | Meaning |
|---|---|
| `none` | content unavailable to that LLM edge |
| `schema` | only strict DescribeData-style metadata |
| `scan` | content must pass configured Shield scanners |
| `raw` | bypass row/value restrictions; baseline secret/PII regex still applies unless explicitly disabled |

```r
shield$register_asset(
  adam_spec,
  name = "adam_spec",
  kind = "spec",
  llm_access = list(prompt = "raw", egress = "scan"),
  scan_secrets = TRUE,
  reason = "Validated public specification",
  expires = "session"
)

prompt_text <- shield$prompt_content("adam_spec")
```

Raw egress never follows from `kind` alone. It requires explicit policy and
provenance:

```r
shield$register_asset(
  adam_spec, name = "adam_spec", kind = "spec",
  llm_access = list(prompt = "raw", egress = "raw"),
  reason = "Validated public specification")

tool_result <- shield$trusted_result(value, source = "adam_spec")
```

An untagged or mixed tool result remains scanned. Raw asset policies require a
reason, expire with the owning DataShield session by default, may set a POSIXct
expiry, and emit bypass audit events. Synthetic raw always keeps baseline
PII/secret scanning; spec raw may explicitly set `scan_secrets = FALSE`.

### Column-level raw access

Assets are whole-object; `register_data(column_access=)` is the column-grained
counterpart for a protected data.frame that contains a few public-dictionary
columns (e.g. an SDTM `TESTCD` codelist) alongside protected ones. It reuses the
same `none`/`schema`/`scan`/`raw` access levels as assets, split into
`prompt`/`egress`, and a raw edge likewise requires a non-empty `reason`.

```r
shield$register_data(
  vs, name = "vs",
  sensitivity   = c(SUBJID = "identifier", TESTCD = "identifier"),
  column_access = list(
    TESTCD = list(prompt = "raw", egress = "raw",
                  reason = "SDTM public codelist", scan_secrets = TRUE)))
```

- `prompt = "raw"` lets `DescribeData` enumerate that column's real values
  (no k-anonymity suppression) so the model can write correct filters.
- `egress = "raw"` removes the column from the value-match index, so its values
  are not withheld from tool output.
- An override **missing its `reason` is a hard error** (`register_data()` refuses
  the dataset), so a mislabeled raw grant cannot pass unnoticed.
  `coverage()$raw_access_columns` counts active overrides.

> **egress tier caveat (kiro round-2 #8 / round-4 #7):** on the EGRESS side only
> `none` (a fixed-string deny of every value) and `raw` (removed from the index)
> have distinct behaviour today; `schema` and `scan` both fall through to the
> ordinary value-match scan — there is no structured schema/scan egress
> difference yet, and column-level `scan_secrets` controls the PROMPT (raw
> DescribeData) path, not egress. Full per-column egress tiers need result→column
> provenance (tracked, not yet implemented). The PROMPT tier implements all four
> levels.

### Host pattern: a provenance-tagging spec tool

Raw asset egress needs a provenance tag. Rather than a framework "trusted tool"
type, a host composes the existing primitives — `register_asset()` plus
`trusted_result()` — inside its own tool:

```r
read_adam_spec_tool <- function(shield) {
  ellmer::tool(
    name = "ReadADaMSpec",
    fun = function(path) {
      text <- readLines(path, warn = FALSE)
      shield$trusted_result(paste(text, collapse = "\n"), source = "adam_spec")
    },
    description = "Read a registered, LLM-safe ADaM specification.",
    arguments = list(path = ellmer::type_string("Spec file path")))
}
# register the asset policy once; the tool auto-tags provenance on every read
shield$register_asset(adam_spec_path, name = "adam_spec", kind = "spec",
  llm_access = list(prompt = "raw", egress = "raw"),
  reason = "Validated public specification")
# separately, register the TOOL itself so the model can call it at all, then
# (re-)install so the egress wrapper picks up trusted_result()'s provenance
chat$register_tool(read_adam_spec_tool(shield))
shield$install(chat)
```

The agent calls `ReadADaMSpec` like any tool; the raw bypass is authorized by
the registered asset policy and audited, and a mixed/untagged result from any
other tool is still scanned.

## P0 — the foundation (available now)

The minimal, deterministic slice that already gives real protection:

```{r}
# Easy entry: strategy specs create one private DataShield R6 for this client.
client <- codeagent_client(chat, data_shield = list(
  shield_describe(k_anon = 5),

  shield_egress(max_rows = 0)
))

# Harness-only client: attach tools, then install its R6 engine.
client <- codeagent_client(chat, register_tools = FALSE,
  data_shield = list(shield_describe(), shield_egress(max_rows = 0)))
chat$register_tool(my_tool)
client$data_shield$install(client$chat)
```

- **Edge 2 — shape-based egress row-cap.** codeagent does **not** inspect code
  or block `print`. It looks only at the **shape** of a tool's returned text:
  output with a data.frame / tibble print signature or a many-row rectangular
  table is truncated to a shape summary; scalars, messages, **model summaries**,
  plots and errors pass through untouched. Content-agnostic and tunable via
  `max_rows`.
- **Edge 1 — ambient stays schema-only.** codeagent's ambient injection already
  emits only `name [data.frame N x M: col:type, ...]` (no values); Data Shield
  keeps it that way.

### Runtime uploads in Shiny

The dataset does **not** need to be known when the app starts. Register it in
the upload observer immediately after reading it; tools may already be attached
and wrapped, because value matching reads the live index at invocation time.

```{r}
# Inside each Shiny server session: one R6 may be shared by selected chats.
shield <- DataShield$new(
  strategies = list(shield_describe(), shield_egress(max_rows = 0)))
data_env <- new.env(parent = emptyenv())

client_factory <- function() {
  codeagent_client(make_chat(), data_shield = shield)
}

observeEvent(input$file, {
  df <- read.csv(input$file$datapath)
  data_env$uploaded <- df
  shield$register_data(df, name = "uploaded") # no advance columns needed
})
```

A complete runnable host-style Shiny example is installed at:

```{r}
system.file("examples/data_shield_upload_app.R", package = "codeagent")
```

From a development checkout:

```r
devtools::load_all(".")
source("inst/examples/data_shield_upload_app.R")
```

It demonstrates five outcomes after upload: bulk rows withheld by `row_cap`, a
single indexed value withheld by `value_match`, unregistered PII redacted by
`shield_regex`, a harmless shape summary passed through, and strict
`DescribeData` metadata with raw identifiers suppressed.

For a smaller single-focus demo — a real `codeagent_client` wired into a real
`shinychat::chat_ui`, with `fileInput()` upload on one side and the live
non-sensitive audit log on the other — see
`inst/examples/data_shield_minimal_app.R`.

> **Multi-user isolation:** create `DataShield$new()` inside the Shiny server
> function, share it only among the intended chat threads, and register data via
> `shield$register_data()`. Other browser sessions receive separate R6 instances
> and cannot see or influence its index.

Illustrative behaviour of the P0 row-cap (implemented):

| tool output | P0 action |
|---|---|
| `print(mtcars)` (bulk rows) | capped → shape summary |
| a tibble print (`# A tibble: 320 x 12`) | capped |
| `print(nrow(df))` → `320` | passes |
| a status message | passes |
| `print(summary(fit))` | passes (not row data) |


## P1 `DescribeData`: strict safe metadata contract

`DescribeData` is the sanctioned way for the model to understand protected data
without receiving rows. Its output is determined by three orthogonal dimensions:

| Dimension | Values | Purpose |
|---|---|---|
| Global policy | `distributions = "off" / "on" / "dp"` | strict defaults to no distributions; `on` is explicit opt-in; `dp` adds noise + budget |
| Column sensitivity | `identifier / quasi / measure / open` | business role that clamps the maximum disclosure for that column |
| Data type | numeric / factor / character / Date / … | determines the safe representation: range, labels, or free-text marker |

Strict (`distributions = "off"`) matrix:

| Metadata | identifier / quasi | measure / open |
|---|---|---|
| column name, type, missing presence | shown | shown |
| numeric/date min–max | hidden | shown |
| low-cardinality categorical labels | hidden | shown without counts; levels with support `< k` are suppressed |
| real free-text examples | hidden | hidden |
| histogram, quantiles, mean/SD, category counts | hidden | hidden (`on`/`dp` opt-in only) |

A factor is not automatically safe: a factorised subject ID remains an
`identifier`. Character columns receive categorical labels only when they are
low-cardinality, low-uniqueness, non-PII, and every exposed level satisfies the
k-anonymity threshold. Free text never receives real examples in strict mode.

### Protected schema in the system prompt

When protected data is registered **before** `codeagent_client()` is built, its
filtered `DescribeData`-style schema is inserted into a `<protected-data>`
system-prompt block. This tells the model the real dataset names, dimensions, and
allowed columns without exposing identifiers or rare categories. `DescribeData`
remains the authoritative live fallback.

`register_data()` updates the live engine and value index but deliberately does
not rewrite an existing Chat prompt. After a runtime upload, refresh explicitly:

```r
shield$register_data(df, name = "uploaded")
client <- refresh_data_shield_context(client)
```

The refresh rebuilds the prompt while preserving history, tools, budgets, hooks,
and the same live `DataShield`; expect a one-time prompt-cache miss. Under
`distributions = "dp"`, both schema-block generation and `DescribeData` consume
the configured per-dataset privacy budget.

## P1.5 ordered egress scanners

Strategy-list order is the execution order. `shield_regex()` catches sensitive
content even when no data.frame was registered:

```r
client <- codeagent_client(chat, data_shield=list(
  shield_egress(max_rows=0),
  shield_regex(on_fail="redact"),
  shield_regex(patterns=c(study_id="STUDY-[0-9]+"),
               include_defaults=FALSE, on_fail="block")
))
```

Built-ins cover email, phone-like strings, common API-token prefixes and
18-character identity-number shapes. `redact` replaces only matched spans;
`block` discards the whole model-facing result. Custom scanner functions may be
appended with `shield$add_scanner(name, fn)`; invalid scanner results/errors fail
closed.


## C2 universal ingress scanning

`shield_ingress()` is installed into codeagent's existing single central
permission gate; it does not create a competing callback/gate. It scans every
tool's arguments before the read-only fast path. A `block` result raises
`tool_reject`; an `ask` result reuses the current CLI/Shiny approval callback,
including tool-call id correlation.

```r
client <- codeagent_client(chat, data_shield=list(
  shield_ingress(on_fail="ask"),
  shield_egress(max_rows=0),
  shield_regex()
))
```

Defaults intentionally do not reject every `Read` or `print`: `nrow(study)` and
`print("done")` pass, while `head(study)`, `dput(study)`, base64/pickle/JSON
serialization, upload-style curl/requests calls, and shell display of data files
are reviewed or blocked.

## C5 portable sandbox and btw boundary

`shield_sandbox()` deliberately preserves coding capability: project and
session-temp default to `rwx`, protected data defaults to `rw` but may be `rwx`,
and process execution stays enabled. Its current portable backend blocks
explicit paths outside allowed roots, rejects symlink escape, and applies
network/process capability policy in the central gate.

btw is not assumed to provide OS isolation. Its file tools enforce cwd with
`fs::path_has_parent()` but a project-internal symlink to an external file passed
our probe; its RunR executes through `evaluate` in the global environment and
only restores cwd/options/envvars. Data Shield therefore applies uniformly to
native, btw, MCP and host tools.

## C4 semantic reviewer

`shield_reviewer()` supplements deterministic ingress rules for indirect aliases,
multi-step serialization and obfuscated source-to-sink intent. Remote reviewers
receive only sanitized code and non-value metadata. A fresh independent Chat is
created per review, with no tools/history. The default factory uses the parent
provider plus `CODEAGENT_FAST_MODEL`; an explicit `client_factory` may provide a
local or specialized reviewer. Missing/failed/invalid reviewers follow
`on_error`; ask falls back to block when no approval channel exists.

## Related data tools and their boundaries

- **`DescribeData`** is registered only when `shield_describe()` is enabled (or
  when the direct constructor uses its default strategies). It returns the
  filtered metadata contract above and sees datasets registered later.
- **`ExploreData`** is a general read-only data-query tool, not a confidentiality
  boundary or OS sandbox. It evaluates supplied R code in a child environment;
  under Data Shield its arguments still pass the central ingress gate and its
  model-facing result still passes egress filtering. Data-frame results expose a
  shape string to the model while rich rows remain a UI artifact; scalar text can
  still be caught by `value_match`/regex.
- **`audit_code_tool(shield, project_root)`** is an opt-in `AuditCode` tool that a
  host may register before executing R code. It parses static references, reads
  only regular source files with allowed extensions inside `project_root`, never
  executes the submitted code, and returns risk/reference metadata rather than
  file contents. With a shield, whitelisted source text goes through the reviewer
  rail. `AuditCode` grants no read/write/shell capability to the main model.

## Roadmap

- **P0.5 — `value_match` (available)**: deterministically catches *targeted*
  leaks the row-cap lets through (e.g. printing one patient's name), by matching
  tool output against high-entropy values registered with
  `shield$register_data()`.
- **P1 — `DescribeData` + protected-data registry (strict available)**: the
  model's sanctioned hardened view (schema, sensitivity, missing presence,
  measure/open ranges and k-supported labels; no distributions/counts/examples
  in strict `"off"` mode). `distributions="on"`/`"dp"` (categorical counts,
  real or DP-noised with a per-dataset budget) are **available**; DP for
  numeric/continuous statistics remains a later phase (needs a host-supplied,
  data-independent clipping bound).
- **P1.5 — ordered scanner pipeline + `shield_regex()` (available)**:
  unregistered PII/secrets are redacted/blocked with precise spans; custom
  scanner failures fail closed.
- **C2 — `shield_ingress()` (available)**: all tool arguments pass the central
  permission gate; deterministic high-confidence rules block or force approval.
- **C5 — portable `shield_sandbox()` (available)**: project/temp `rwx`, protected
  data `rw` by default, realpath/symlink containment and policy fallback; full OS
  process adapter remains roadmap.
- **C4 — `shield_reviewer()` (available)**: optional sanitized ingress-code
  semantic rail using a fresh small-model Chat; remote raw output remains forbidden.
- **P2 — full OS sandbox adapter and differential privacy for numeric/continuous
  statistics (opt-in).**

## Sub-agent boundary

Foreground sub-agents (`Agent`) inherit the exact same live `DataShield` R6;
their tools are installed behind the same central permission and Shield gates
before the first child request. This applies to synchronous and concurrent async
Agent calls. While a shield is active, raw `btw_tool_agent_*`/custom delegation
paths that cannot accept the engine are not exposed. The child reply is output-
gated before `SubagentStop` hooks, callbacks, or the parent tool result, and
shielded sidechains are not persisted.

`BackgroundAgent` and `/bg` currently **fail closed** under Data Shield: their
mirai worker is a separate R process and cannot safely share the session's R6
state or protected-value index. Use foreground `Agent` until a per-owner worker
reconstruction protocol is implemented.


## Combination safety: what each combination actually protects against

Data Shield is composed from independent strategies, so it is possible to
enable a combination that *looks* protective but is not. The table below is
sourced directly from `tests/testthat/test-data-shield-combinations.R` (a CI
suite, not prose): a future refactor that breaks any of these conclusions
fails that suite immediately, rather than silently drifting out of date.

| Combination | Bulk dump | Targeted single value | Alias bypass (`y <- study; print(y)`) | Verdict |
|---|:---:|:---:|:---:|---|
| `shield_egress()` alone | withheld | withheld | withheld (egress inspects output content, not the code path) | ✅ safe floor |
| `egress` + `ingress` + `regex` (recommended) | withheld | withheld | withheld | ✅ recommended |
| **`shield_ingress()` alone** | **leaks** | **leaks** | **leaks (confirmed)** | ⚠️ **not safe alone** |
| **`shield_describe()` alone** | **leaks** | **leaks** | — | ⚠️ **not safe alone** (only governs the model's own metadata query; does not filter other tools' output) |
| `shield_regex()` alone | — | leaks for non-PII-shaped custom ids | — | ⚠️ only catches common PII shapes |

**The one-line takeaway: `shield_egress()` is the only non-optional boundary.
Every other strategy is defense-in-depth and cannot substitute for it.**

### Three ready-to-use combination templates

`shield_preset_strict()`, `shield_preset_balanced()`, and
`shield_preset_clinical()` are callable functions returning these exact
combinations -- no copy-pasting required:

```r
# Strict: compliance / audit demos
strict <- shield_preset_strict()
# shield_describe(k_anon = 5),
# shield_egress(detectors = c("row_cap", "value_match"), max_rows = 0, on_fail = "block"),
# shield_regex(on_fail = "block"),
# shield_ingress(on_fail = "block")

# Balanced: everyday development, low friction
balanced <- shield_preset_balanced()
# shield_egress(max_rows = 0, on_fail = "redact"),
# shield_regex(on_fail = "redact")

# Clinical: adds the semantic reviewer + strict k-anonymity
clinical <- shield_preset_clinical()
# shield_describe(k_anon = 5),
# shield_egress(max_rows = 0),
# shield_regex(),
# shield_ingress(on_fail = "ask"),
# shield_reviewer(model = Sys.getenv("CODEAGENT_FAST_MODEL"), on_risk = "ask")

client <- codeagent_client(chat, data_shield = shield_preset_strict())
```


### Two intentionally UNSAFE demo combinations

These reproduce the two "not safe alone" rows of the matrix above, exactly as
wired in `inst/examples/data_shield_minimal_app.R`'s "Shield strength"
selector — switch to one there, ask the chat to dump the uploaded data, and
watch it leak live.

```r
# UNSAFE: ingress-only — no egress boundary at all.
unsafe_ingress_only <- list(shield_ingress(on_fail = "block"))

# UNSAFE: describe-only — DescribeData is registered, but nothing filters
# what OTHER tools return.
unsafe_describe_only <- list(shield_describe(k_anon = 3))
```

- **`unsafe_ingress_only`** only scans tool *arguments* before execution (see
  `shield_ingress()` in the parameter reference above). It has no `shield_egress()`,
  so nothing inspects what a tool actually *returns*. Its static regex rules
  also only see the literal code text: `y <- study; print(y)` does not match
  any "print/dump a known dataset name" pattern, so the alias sails through —
  this is the exact bypass the combination-safety table above marks "confirmed".
- **`unsafe_describe_only`** only registers the `DescribeData` tool (see
  `shield_describe()` above). That tool is the model's own sanctioned query
  path; it does nothing to any *other* tool's return value, so a tool that
  simply returns the raw data.frame is never touched.
- Both omit `shield_egress()` — see "Combination safety" above for why that
  is the one strategy every other combination depends on.

`inst/examples/data_shield_minimal_app.R` has a live "Shield strength"
selector covering the three templates above plus these two intentionally
UNSAFE combos, so you can switch between all five in a running chat and watch
the exact leaks in the table happen live.

## Honest limits

Data Shield **reduces** disclosure risk; it does not eliminate it. Deterministic
detectors (row-cap, `value_match`, regex) miss adversarially obfuscated egress
(e.g. base64-encoding data before printing); those are mitigated — not solved —
by the ingress blacklist and the no-network sandbox. The strongest guarantees
come from the structural layers (metadata-only feeding + no-network execution),
with scanning as defense-in-depth.

Specific residual risks to weigh before relying on it:

- **Layer composition matters — see the combination-safety table above.**
  Enabling only `shield_ingress()` or only `shield_describe()` without
  `shield_egress()` is **not** a safe configuration. Do not omit egress.
- **The semantic reviewer is itself an LLM.** `shield_reviewer()` can be evaded
  by a sufficiently obfuscated source-to-sink path and adds latency/cost per
  reviewed call; it is defense-in-depth over the deterministic rails, not a
  guarantee.
- **`value_match` scales linearly and is now bounded.** Benchmarked on
  open-source CDISC-ADaM-format example data from the {pharmaverse} project
  (`inst/bench/value_match_benchmark.R`): ~130 MB of keys and ~10 s to
  index 1M high-entropy values, with zero false positives on ordinary clinical
  prose and pharmaverse-format `USUBJID`/`SUBJID` caught. Because memory grows
  linearly, `register_data(max_index_values=)` caps the index (default 500 000,
  ~65 MB). If the cap would truncate the index, registration errors and refuses
  the partially indexed dataset; raise the cap, split the data, or explicitly use
  `NULL`/`Inf`. The `min_len`/`min_card` thresholds performed well on these ids
  and are unchanged.
- **Image tool results and full OS isolation remain roadmap** (see the status
  banner): a tool-returned rendered table/plot of raw rows bypasses text egress
  scanning, and the portable sandbox is a path/capability policy, not kernel
  isolation. Prompt-image OCR is a separate opt-in edge-1 control described
  above; its optional-dependency fallback is intentionally a documented blind
  spot.
- **`shinychat` text-bearing attachments ARE scanned at edge 1.** codeagent's
  main UI enables `chat_ui(allow_attachments = TRUE)`; the input gate extracts
  and scans text-bearing attachments (e.g. `ContentPDF`), failing **closed** to
  block when the content cannot be verified (immutable Content cannot be redacted
  in place). Image attachments remain a blind spot unless an OCR scanner
  (`data_shield_ocr_scanner()`) is wired in — with a scanner configured, an OCR
  or scan failure fails closed to block. `fileInput()` → `register_data()`
  remains the controlled, indexed path for tabular uploads.
- **Data Shield does not govern destructive operations** (`rm -rf`, dropping a
  table, force-pushing) — that is a different axis (operation safety, not
  data confidentiality) covered by the permission gate and hooks, not by any
  `shield_*()` strategy here. `shield_ingress()`'s pattern-matching pipeline
  *can* be repurposed for it, but its built-in defaults are exfiltration-
  focused, not destructiveness-focused. See `vignette("permissions")`'s
  "Hooks" section for the actual mechanisms (a `rules` deny glob, a
  `PreToolUse` hook, or a custom `shield_ingress()` pattern) and why none of
  them are foolproof against rewording.


