---
title: "Getting started with codeagent"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with codeagent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`codeagent` is an R-native coding agent: it wraps any `ellmer` chat with a full
agent harness (tools, permissions, compaction, hooks, skills) and drives it from
the R console, a terminal REPL, or a Shiny app. This vignette walks from a first
query to configuration and sandboxing.

All code chunks below require a configured model endpoint and are therefore not
evaluated when the vignette is built.

## 1. Build a client

A `codeagent` client wraps an `ellmer::Chat`. Bring your own backend --
Databricks, Anthropic, OpenAI-compatible, or Ollama:

```{r}
library(codeagent)

chat <- ellmer::chat_openai_compatible(
  base_url    = Sys.getenv("CODEAGENT_BASE_URL"),
  model       = Sys.getenv("CODEAGENT_MODEL"),
  credentials = function() Sys.getenv("CODEAGENT_API_KEY")
)

client <- codeagent_client(chat, permission_mode = "bypass")
```

`codeagent_client()` injects the tool set (Bash, Read, Write, Edit, Glob, Grep,
LS, plus `btw` tool groups) and builds the system prompt. The `permission_mode`
controls how tool calls are gated -- see section 5.

## 2. One-shot queries

```{r}
codeagent(client, "List the .R files in R/ and summarise what each does")
```

The agent plans, calls tools, and returns a final answer. History accumulates on
the client, so follow-up calls keep context.

## 3. Interactive REPL

From R:

```{r}
codeagent_repl(client)
```

Or install the command-line executable and use it from a terminal:

```{r}
install_codeagent_cli()
```

```bash
codeagent chat          # interactive REPL
codeagent run "..."     # one-shot
codeagent app           # launch the Shiny UI
```

Inside the REPL, slash commands include `/model`, `/compact`, `/clear`,
`/rewind`, `/sessions`, and `/budget`; `/name` invokes a skill.

## 4. Shiny app

```{r}
codeagent_app(client, theme = "default")
```

The app streams output, renders tool cards, and provides session management and
a searchable skill panel.

## 5. Permissions

Permission mode decides whether a tool call is allowed, denied, or requires
confirmation:

| Mode | Behaviour |
|------|-----------|
| `default` | reads auto-allow; writes and shell require confirmation |
| `plan` | read-only; all mutating tools denied |
| `accept_edits` | file edits auto-allow; Bash still asks |
| `bypass` | everything allowed |
| `dont_ask` | mutating tools auto-denied (CI) |
| `bubble` | sub-agent mode; decisions bubble to the parent |

Fine-grained rules match on tool arguments -- for example, allow only a specific
command:

```{r}
client <- codeagent_client(
  chat,
  permission_mode = "default",
  rules = list(PermissionRule("Bash", "allow", rule_content = "R CMD build *"))
)
```

## 6. Configuration with settings.json

Scaffold a settings file:

```{r}
use_codeagent_settings(scope = "user")
```

It mirrors the schema of command-line coding agents (model tiers, an `env`
block, permissions, hooks, MCP servers, sandbox, effort level). Precedence, low
to high: package defaults, then `~/.codeagent/settings.json`, then
`.codeagent/settings.json`, then environment variables.

Keep API keys in `.Renviron` (as `CODEAGENT_API_KEY`), never in
`settings.json`.

## 7. Sandboxed R execution

The `RunR` tool executes R code behind the permission gate. Enable sandboxing to
run each call in an isolated `callr` subprocess with a scrubbed environment (API
keys are not visible), no `.Renviron` reload, and a wall-clock timeout:

```json
{ "sandbox": { "enabled": true, "allow_network": false } }
```

Sandboxing is opt-in because spawning a subprocess has a small cost. Turn it on
when executing less-trusted code; leave it off for fast, local, trusted use.

## 8. Multi-agent teams

Run independent tasks in parallel, capped to the container's CPU quota:

```{r}
# Fixed fan-out
team_run(c("review R/a.R", "review R/b.R"))

# Work-stealing over a shared board (balances uneven task sizes)
team_coordinate(c("task 1", "task 2", "task 3", "task 4"))
```

## Where to go next

* `?codeagent_client` -- all client options (tool groups, verification, MCP).
* `?codeagent_app` -- Shiny app themes and panels.
* `README` -- feature overview and architecture.
