---
title: "Permissions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permissions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, purl=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

**Language:** English | [简体中文](permissions-cn.html)

Every registered tool call passes through one central authorization gate before
it runs. Native, btw, Format, and MCP tools are built in bypass mode; the gate
installed on `chat$on_tool_request` is the sole permission authority. Gate
installation fails closed if that callback cannot be registered.

## How the gate decides

```
model requests a tool
    |
    v
chat$on_tool_request -> central gate (.tool_gate_fn)
    |
    +- Data Shield ingress scan (when a shield is active; every tool)
    |     block/error -> PermissionDenied -> ellmer::tool_reject()
    |
    +- capability == read AND no per-tool override AND no Shield ask?
    |     yes -> ALLOW at the gate (fast path)
    |
    v  .gate_decide() (highest precedence first)
  1. settings$tools$overrides[tool]      -> allow / deny / ask
  2. settings$tools$capabilities[class]  -> write|exec|net -> allow/deny/ask
  3. fallback -> check_permission(mode, rules):
       plan          -> deny non-read before checking rules
       user rules    -> first matching glob wins
       accept_edits  -> edit tools allow
       bypass        -> allow
       bubble        -> ask (parent/host resolves it)
       dont_ask      -> read-only allow, otherwise deny
       auto          -> fast-model classifier -> allow/deny/ask
       default       -> read-only and recognized read-only Bash allow;
                        otherwise ask
    |
    v
  decision --+-- deny -> PermissionDenied -> ellmer::tool_reject()
             +-- ask  -> ask_fn(): CLI console prompt or async Shiny bar
             |            reject/error/no ask_fn -> deny
             |            approve -> continue
             +-- allow ---------------------------> continue
                                                    |
                                                    v
                         PreToolUse wrapper: deny or rewrite arguments;
                         rewritten arguments are rechecked by gate/Shield
                                                    |
                                                    v
                         tool runs -> on_tool_result -> PostToolUse hook
```

The read fast path is deliberate: a read-capability tool with no explicit
per-tool override is allowed after the Shield scan, before capability policy,
rules, or mode fallback. Use `settings$tools$overrides` when a read tool must be
asked about or denied. Unknown host tools default to the `read` capability for
compatibility, so hosts should declare them with `register_tool_meta()` or
`tool_meta` when installing the gate.

`settings$tools$sets` (`"A"` = codeagent core, `"B"` = btw) controls which tool
sets are registered. It is registration policy, not a per-call decision.

## Modes

These are fallback behaviors. A per-tool override or non-read capability policy
can supersede them.

| Mode | Fallback behavior |
|------|-------------------|
| `default` | Read-only tools and recognized read-only Bash commands are allowed; other calls ask |
| `plan` | Non-read calls are denied before user rules; read tools are allowed |
| `accept_edits` | File edit tools are allowed; other non-read calls still ask |
| `bypass` | All calls are allowed (use with care) |
| `dont_ask` | Read-only calls are allowed; non-read calls are denied, which suits unattended runs |
| `auto` | The configured fast model classifies the call as allow, deny, or ask |
| `bubble` | Returns ask so a parent agent or host approval callback can decide |

```{r, purl=FALSE}
client <- codeagent_client(chat, permission_mode = "default")
```

## Fine-grained rules

A `PermissionRule` first glob-matches the tool name, then optionally matches the
relevant argument: `command` for Bash, `file_path` for Read/Write/Edit/MultiEdit,
and `pattern` for Glob/Grep. Matching is case-sensitive and the first match
wins. Rules supplied directly to `codeagent_client(rules=)` precede rules loaded
from settings.

Settings arrays are converted in `allow`, then `deny`, then `ask` order. Avoid
overlapping patterns: a deny entry does not automatically outrank an earlier
allow entry. Rules are reached only after per-tool and capability policy, and
not by the read fast path described above.

```json
{
  "permissions": {
    "allow": ["Bash(git status)", "Read(*)"],
    "deny":  ["Bash(rm -rf *)"],
    "ask":   ["Write(*)"],
    "defaultMode": "default"
  }
}
```

## Public signatures and defaults

```{r, purl=FALSE}
PermissionRule(
  tool_name,
  behavior = c("allow", "deny", "ask"),
  source = "session",
  rule_content = NULL
)

check_permission(
  tool_name,
  mode = "default",
  rules = list(),
  tool_input = NULL
)

install_permission_gate(
  chat,
  permission_mode = "default",
  rules = list(),
  tools = list(),
  ask_fn = NULL,
  tool_meta = list()
)

codeagent_client(
  chat = NULL,
  permission_mode = "default",
  rules = list(),
  cwd = getwd(),
  max_turns = 100L,
  btw_groups = NULL,
  worktree_isolation = FALSE,
  verify_fn = NULL,
  mcp_config = NULL,
  register_tools = TRUE,
  data_shield = NULL,
  max_budget_usd = NULL
)
```

`install_permission_gate()` is for attaching the central gate to an existing
`ellmer::Chat`. `tools` has the same `sets` / `capabilities` / `overrides` shape
as `settings$tools`; `tool_meta` is a named tool-to-capability list. In normal
codeagent use, `codeagent_client()` loads settings and installs the gate.

## Interactive approval

In `default` mode, the Shiny app displays an Allow/Deny bar above the composer
when a call resolves to `ask`; execution resumes when the promise resolves. The
CLI uses a synchronous console prompt. Failure, rejection, or a missing
`ask_fn` is denial. `AskUserQuestion` uses a separate async question bar to
pause for a clarifying answer.

## Hooks and rewritten input

`PreToolUse` does not make the gate's initial decision. It runs once in the tool
wrapper after gate authorization (and therefore after any human approval), but
before execution. It may deny the call or return `updatedInput`; rewritten
arguments are checked again against the live permission policy and Data Shield,
and an `ask` result with no second approval path is denied. `PermissionDenied`
fires on gate denial, and `PostToolUse` fires from `on_tool_result` after
execution.

This ordering means a `PreToolUse` veto still prevents execution, but it does
not prevent an approval prompt from first being shown.

## Dangerous approvals and defense in depth

If an operator approves `Bash: rm -rf ./data`, approval alone does not establish
that the operation is safe. Available veto points have different timing:

| Layer | Timing and limitation |
|---|---|
| `settings$tools$overrides` deny | Denies before prompting; use this for unconditional per-tool policy, including read tools |
| A deny `PermissionRule` | Denies before prompting when fallback rule evaluation is reached; it is bypassed by stronger policy and the read fast path |
| Data Shield ingress/tool policy with a blocking result | Runs before the read fast path and before prompting |
| `PreToolUse` returning deny | Runs after gate/human authorization but still before tool execution |

These controls are syntactic or policy-based defense in depth. For example,
`Bash(rm -rf *)` does not catch equivalent behavior routed through `RunR`,
`unlink(..., recursive = TRUE)`, another destructive command, runtime string
construction, or a script written in one call and executed in another.

A semantic reviewer of operation intent could raise the bar but would remain
model-based. A full OS-level sandbox with read-only or scoped-writable mounts
would enforce the boundary independently of command spelling; the current
portable `shield_sandbox()` policy backend is not such a kernel-enforced
adapter. A curated cross-language destructive-operation strategy remains an
open design question, not an implemented guarantee.

