---
title: "Skills and slash commands"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Skills and slash commands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What are skills?

Skills are reusable prompt templates that the agent can invoke on demand.
Type `/skillname` in the chat to activate a skill, or the agent will
auto-trigger matching skills based on your request.

## Built-in skills

codeagent ships with skills for common R package development tasks:

| Skill | Description |
|-------|-------------|
| `/plan` | Enter read-only planning mode |
| `/verify` | Run tests and check the package |
| `/compact` | Compact the context window |
| `/roxygen` | Generate roxygen2 documentation |
| `/testthat` | Write unit tests |
| `/document` | Run `devtools::document()` |
| `/lint` | Style and lint R code |
| `/news` | Update NEWS.md |
| `/explore` | Explore a data.frame |

## Using skills from the chat

```
# In the Shiny app or REPL:
/plan add a new summarise_by_group() function
/roxygen summarise_by_group
/testthat summarise_by_group
```

## Creating custom skills

Skills are `name/SKILL.md` directories in any of the search paths:

```
~/.codeagent/skills/my-skill/SKILL.md
.codeagent/skills/my-skill/SKILL.md    # project-local
.btw/skills/my-skill/SKILL.md
.claude/skills/my-skill/SKILL.md
```

**SKILL.md format**:

```markdown
---
name: my-skill
description: Short description shown in skill picker
argument-hint: "<what to type after /my-skill>"
allowed-tools:
  - Read
  - Bash
---

Skill body — instructions for the agent.

Use $ARGUMENTS to insert the user-supplied arguments.
```

## Skill discovery

```{r}
# List all available skills
list_skills_meta()

# Load a skill prompt programmatically
load_skill_prompt("plan", args = "add feature X", cwd = getwd())
```
