---
title: "Multi-agent team coordination"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-agent team coordination}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, purl=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Language:** English | [简体中文](team-coordination-cn.html)

## Overview

codeagent provides several distinct ways to delegate work. Choose one according
to whether tasks are independent, dependency-constrained, foreground, or truly
non-blocking:

- `team_run()` is fixed fan-out for independent tasks. It runs workers in
  parallel but waits for every result before returning.
- `team_coordinate()` is a blocking work-stealing pool backed by a shared
  SQLite task board. It supports a dependency DAG and uneven task durations.
- `team_lead()` asks an LLM lead to decompose, review, and optionally re-plan a
  goal over several blocking `team_coordinate()` rounds.
- The foreground `Agent` tool delegates one subtask. On an async parent turn,
  opt-in async sub-agents can run multiple `Agent` calls concurrently.
- `BackgroundAgent` and `/bg` are fire-and-forget paths whose results are
  surfaced on a later turn.

`team_run()` and `team_coordinate()` require `mirai`. Board operations require
`DBI` and `RSQLite`. Their worker permission mode defaults to `"bypass"`
because daemon processes cannot answer interactive approval prompts; use that
default only for tasks and environments you trust.

## Coordination at a glance

```
team_run(tasks)                       FIXED FAN-OUT (blocking caller)
  N mirai daemons; one codeagent query per task
  -> wait for all tasks -> list of results in input order

team_coordinate(tasks, blocked_by)   WORK-STEALING (blocking caller)
  board_create(): tasks + deps + messages tables in SQLite
  seed all tasks -> wire blocked_by indices -> reject dependency cycles
  N workers, each: repeat {
    board_claim() -- atomic claim of lowest-id eligible pending task
      | none claimable
      +- no unfinished tasks       -> break
      +- reclaim stale claims      -> retry abandoned work
      +- board is truly stalled    -> break
      +- blocker still running     -> sleep(backoff), then retry
    run codeagent(task)
    board_complete(result)
    board_send_message("completed task", recipient = "coordinator")
  }
  -> wait for workers -> final board data.frame

team_lead(goal, max_rounds)           LLM-LEAD LOOP (blocking caller)
  structured decomposition -> tasks + dependency DAG
  -> team_coordinate(...)
  -> structured review: done or follow-up tasks
  -> repeat with a new board until done, no tasks, or max_rounds
```

## Fixed fan-out: `team_run()`

`team_run(tasks, model = NULL, n_workers = NULL, permission_mode = "bypass",
cwd = getwd())` creates a fresh client for every task and returns a list in the
same order as `tasks`. A failed task is represented by an `"[Error] ..."`
string rather than aborting the whole result collection.

```{r, purl=FALSE}
library(codeagent)

# Review multiple files in parallel
results <- team_run(c(
  "Review R/tool_display.R for any issues",
  "Review R/permissions.R for any issues",
  "Review R/compaction.R for any issues"
))

# Each element of results is the agent's response for that task
cat(results[[1]])
```

Worker count defaults to
`min(length(tasks), parallelly::availableCores())`, with a conservative fallback
when `parallelly` is unavailable. An explicit `n_workers` is capped at the same
cgroup-aware limit. `team_run()` has no worktree argument, so tasks that write
to the same checkout can collide; reserve it for independent read-only work or
otherwise isolated resources.

## Work-stealing: `team_coordinate()`

The current signature is:

```r
team_coordinate(
  tasks,
  model = NULL,
  n_workers = NULL,
  permission_mode = "bypass",
  cwd = getwd(),
  blocked_by = NULL,
  worktree = FALSE,
  backoff = 0.5,
  reclaim_timeout = 300,
  db_path = tempfile(fileext = ".sqlite")
)
```

Workers repeatedly claim work, so a faster worker can process more than one
item. The function itself waits for all worker loops before returning the final
board with columns `id`, `prompt`, `owner`, `status`, and `result`.

```{r, purl=FALSE}
results_df <- team_coordinate(
  tasks = c("task 1", "task 2", "task 3", "task 4", "task 5"),
  n_workers = 2
)
# Returns a data.frame with columns: id, prompt, owner, status, result
print(results_df[, c("prompt", "status", "owner")])
```

### Dependencies

`blocked_by[[i]]` contains 1-based indices into `tasks`. A task is claimable only
when every valid blocker is `done`. Cycles are rejected before workers launch;
out-of-range indices and self-dependencies are ignored.

```{r, purl=FALSE}
dag_result <- team_coordinate(
  tasks = c(
    "Create the schema",
    "Implement against the schema",
    "Run integration checks"
  ),
  blocked_by = list(integer(0), 1L, 2L),
  n_workers = 2
)
```

Claims use a `BEGIN IMMEDIATE` SQLite transaction and select the lowest-id
eligible task, so dependency-free boards are FIFO. If no task is claimable,
workers distinguish a completed board, an in-progress blocker, and a true
stall. Idle live workers reclaim claims older than `reclaim_timeout`; if every
worker process has died, no process remains to perform that reclamation.

### Worktree isolation

Set `worktree = TRUE` to request one temporary detached Git worktree per worker:

```{r, purl=FALSE}
review <- team_coordinate(
  c("Inspect module A", "Inspect module B"),
  n_workers = 2,
  worktree = TRUE
)
```

This is best-effort: when Git or a repository worktree is unavailable, the
worker falls back to `cwd`. Temporary worktrees are force-removed when workers
exit. codeagent does not merge or copy edits back, so use this mode for isolated
analysis unless another explicit workflow preserves the changes.

## The shared task board

`board_create(db_path = tempfile(fileext = ".sqlite"))` creates `tasks`, `deps`,
and `messages` tables and invisibly returns the database path. `board_claim()`
is atomic across processes. `board_complete()` records a result, and
`board_status()` returns the current task table.

### Inter-agent messaging

The shared board also supports broadcast and directed message records:

```{r, purl=FALSE}
db <- board_create()
board_add_task(db, "analyse the sales data")
board_add_task(db, "generate the summary report")

# Worker 1 claims a task
task <- board_claim(db, worker_id = "w1")
board_send_message(db, sender = "w1", body = "Starting analysis...",
                   recipient = "coordinator")

# Complete with result
board_complete(db, task$id, result = "Analysis complete: 3 trends found")
```

`recipient = NULL` creates a broadcast. `board_messages(db, "coordinator")`
returns broadcasts plus messages addressed to that recipient, while
`board_messages(db)` returns the entire log. These messages are not
automatically inserted into worker prompts: the built-in coordinator writes a
completion notice but does not use the message table as an agent conversation
channel.

### Watching and displaying a board

`board_watch(db, callback, latency = 0.3)` starts an event-driven file watcher
when the optional `watcher` package is available and otherwise returns `NULL`.
`team_dashboard(db_path, poll_ms = 1500L)` uses that watcher and falls back to
polling. To observe a live `team_coordinate()` run, pass a stable `db_path` and
run the blocking coordinator in another process or background job.

```{r, purl=FALSE}
db <- tempfile(fileext = ".sqlite")
# In another R process:
# team_coordinate(c("A", "B", "C"), db_path = db)
team_dashboard(db)
```

## LLM-led rounds: `team_lead()`

`team_lead(goal, model = NULL, cwd = getwd(), max_rounds = 3L,
n_workers = NULL, permission_mode = "bypass", worktree = FALSE,
decompose_fn = NULL, review_fn = NULL, coordinate_fn = NULL)` first asks the
lead model for structured tasks and dependencies. The three callback arguments
are injectable seams primarily used for testing or custom orchestration. After each coordinated
round it asks whether the goal is complete and, if needed, runs only the
follow-up plan. Non-empty executed boards are row-bound with a `round` column;
if decomposition produces no tasks before any round, the empty return has only
the base board columns. `max_rounds` is coerced to at least one. A review error is handled as “done,” so
it stops rather than retrying indefinitely.

```{r, purl=FALSE}
lead_result <- team_lead(
  "Review the parser, fix confirmed defects, and verify the focused tests",
  max_rounds = 3,
  n_workers = 2,
  worktree = TRUE
)
```

## Foreground, concurrent, and background sub-agents

The agent can invoke the foreground `Agent` tool from chat. The normal path is
synchronous. Worktree isolation is configured on the client and forces
codeagent's owned Agent implementation, which creates and cleans a temporary
worktree:

```{r, purl=FALSE}
client <- codeagent_client(chat,
  permission_mode = "bypass",
  worktree_isolation = TRUE   # each sub-agent in its own git worktree
)
```

As with team worktrees, this is best-effort and does not merge edits into the
main checkout.

`team_run()` and `team_coordinate()` use parallel workers internally but remain
blocking APIs. A different opt-in setting, `async_subagents`, lets multiple
foreground `Agent` tool calls return promises and execute concurrently only
when the parent is already running through an async streaming turn. Sync
one-shot/console turns fall back to the synchronous Agent path.

```json
{
  "async_subagents": true,
  "background_agents": true
}
```

`background_agents = true` registers the model-facing `BackgroundAgent` tool.
It returns a task id immediately; an in-memory registry polls the dedicated
`mirai` compute profile and injects each completed result into a later system
reminder once. The `/bg <task>` and `/bgstatus` local commands expose the same
background registry. Background agents are disabled while Data Shield is
active because a separate process cannot safely inherit the live protected-data
index; use the foreground `Agent` tool in that case.
