---
title: "Multi-agent team coordination"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-agent team coordination}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, purl=FALSE, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

codeagent supports running multiple independent sub-agents in parallel for
tasks that can be divided and executed concurrently.

## Fixed fan-out: team_run()

`team_run()` assigns one task per worker and collects all results:

```{r, purl=FALSE, purl=FALSE}
library(codeagent)

# Review multiple files in parallel
results <- team_run(c(
  "Review R/tool_display.R for any issues",
  "Review R/permissions.R for any issues",
  "Review R/compaction.R for any issues"
))

# Each element of results is the agent's response for that task
cat(results[[1]])
```

Worker count defaults to `min(#tasks, parallelly::availableCores())`
to respect container CPU limits.

## Work-stealing: team_coordinate()

`team_coordinate()` uses a shared SQLite task board where workers
claim tasks dynamically -- faster workers take more tasks:

```{r, purl=FALSE, purl=FALSE}
results_df <- team_coordinate(
  tasks = c("task 1", "task 2", "task 3", "task 4", "task 5"),
  n_workers = 2
)
# Returns a data.frame with columns: id, prompt, owner, status, result
print(results_df[, c("prompt", "status", "owner")])
```

### Inter-agent messaging

The shared board also supports messages between agents:

```{r, purl=FALSE, purl=FALSE}
db <- board_create()
board_add_task(db, "analyse the sales data")
board_add_task(db, "generate the summary report")

# Worker 1 claims a task
task <- board_claim(db, worker_id = "w1")
board_send_message(db, sender = "w1", body = "Starting analysis...",
                   recipient = "coordinator")

# Complete with result
board_complete(db, task$id, result = "Analysis complete: 3 trends found")
```

## Sub-agents in the Shiny app

The agent can invoke sub-agents via the `Agent` tool directly from
the chat. Enable it by registering the agent tool:

```{r, purl=FALSE, purl=FALSE}
client <- codeagent_client(chat,
  permission_mode = "bypass",
  worktree_isolation = TRUE   # each sub-agent in its own git worktree
)
```
